### 2015-03-10 docs/paper.tex
A bit of paper

\documentclass{sigplanconf}
\usepackage{amsmath}
\usepackage{fancyvrb}

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{2015} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

\title{Title Text}

\authorinfo{Iavor S. Diatchki}
           {Galois Inc.}
           {iavor.diatchki@gmail.com}

\maketitle

\begin{abstract}
We present a technique for integrating GHC's type-checker with the
functionality provided by an SMT solver.  The technique was developed
to add support for reasoning about type-level functions on natural
numbers, and so our implementation mainly uses the theory of linear
arithmetic.  The technique, however, is more general and makes it
possible to experiment with other decision procedures supported by
the SMT solver.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\section{Introduction}

\section{Background}

\subsection {SMT Solvers}

SMT solvers, such as CVC4 \cite{cvc4}, Yices \cite{yices}, and Z3
\cite{z3}, implement a wide collection of decision procedures,
which have ...

From a user's perspective, the core functionality of an SMT solver is fairly
simple: we may declare uninterpreted constants, assert formulas,
and check if the asserted formulas are {\em satisfiable}.  Checking
for satisfiability simply means that we are asking the
question: are there concrete values for the uninterpreted constants
that make all asserted formulas true.  Here is an example, using
the notation of the SMTLIB standard \cite{smtlib}

\begin{Verbatim}
(declare-fun x () Int)
(assert (>= x 0))
(assert (= (+ 3 x) 8))
(check-sat)
\end{Verbatim}

The example declares a constant, $x$, asserts some formulas
about it, and then asks the solver if the asserted formulas are
satisfiable.  In this case, the answer is affirmative, as choosing
$5$ for $x$ will make all formulas true.  Indeed, if an
SMT solver reports that a set of formulas is satisfiable, typically
it will also provide a {\em satisfying assignment}, which maps
the uninterpreted constants to concrete values that make the
asserted formulas true.

It is common to use the same machinery for proving the validity of
universally quantified formulas, by looking for counter-examples.
For example, if we want to prove that $\forall x. (3 + x = 8) \implies x = 5$,
then we can use the SMT solver to try to find some $x$ that contradicts
the formula:

\begin{Verbatim}
(declare-fun x () Int)
(assert (= 3 x) 8)
(assert (not (= x 5)))
(check-sat)
\end{Verbatim}

To invalidate an implication, we need to assume the premise, and try to
invalidate the conclusion, which is why the second assertion in the
example is negated. In this case the SMT solver will respond that the
asserted formulas are not satisfiable, which means that there are no
counter examples to the original formula, and so we can conclude that
it must be valid.

\section{}

  * Implication constraints
  * Consistency of assumptions
  * Consistency

  * Relation to Selson Oppen

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}

\end{document}

### 2015-03-10 docs/paper.tex
Checkpoint.

@@
 \usepackage{fancyvrb}
+\usepackage{url}
 
@@
 
+For a few years now, there has been a steady push in the Haskell
+community to explore and extend the type system, slowly approximating
+functionality available in dependently typed languages
+\cite{Eisenberg2012,Lindley2013,Eisenberg2014}.  The additional
+expressiveness enables Haskell programmers to express more
+invariants at compile time, which makes it easier to develop
+reliable software, and also makes Haskell a nice language
+for embedding domain specific languages \cite{ivory-experience}.
+The Haskell compiler is not just a translator, but it also
+serves as a tool that analyzes the program, and helps
+finds common mistakes early in the development cycle.
+
+Unfortunately, the extra expressiveness comes at a cost:
+Haskell's type system is by no means simple. Writing
+programs that make use of invariants encoded in the
+type system can be complex and time consuming.  For
+example, often one spends a lot of time proving simple theorems
+about arithmetic which, while important to convince the compiler
+that various invariants are preserved, contribute little to
+the clarity of the algorithm being implemented \cite{Lindley2013}.
+
+Given that in many cases the proofs constructed by the programmers
+tend to be fairly simple, we can't help but wonder if there might
+be a away to automate them away, thus gaining the benefits of
+static checking, but without cluttering the program with trivial
+facts about, say, arithmetic.  This paper presents a technique
+to help with this problem: we show one method of integrating
+an SMT solver with GHC's type checker.  While we present
+the technique in the context of Haskell and GHC, the technique
+should be applicable to other programming languages and compilers too.
+
 \section{Background}
@@
 SMT solvers, such as CVC4 \cite{cvc4}, Yices \cite{yices}, and Z3
-\cite{z3}, implement a wide collection of decision procedures,
-which have ...
+\cite{z3}, implement a wide collection of decision procedures that
+work together to solve a common problem.  They have proved a useful
+tool in both software and hardware verification.  In this Section,
+we present a very brief introduction to the core functionality
+of a typical SMT solver---readers familiar with the tools could
+skip this Section.
 
@@
 that make all asserted formulas true.  Here is an example, using
-the notation of the SMTLIB standard \cite{smtlib}
+the notation of the SMTLIB standard \cite{smtlib2}
 
@@
 It is common to use the same machinery for proving the validity of
-universally quantified formulas, by looking for counter-examples.
+universally quantified formulas by looking for counter-examples.
 For example, if we want to prove that $\forall x. (3 + x = 8) \implies x = 5$,
@@
 invalidate the conclusion, which is why the second assertion in the
-example is negated. In this case the SMT solver will respond that the
+example is negated. In this case the SMT solver will tell us that the
 asserted formulas are not satisfiable, which means that there are no
-counter examples to the original formula, and so we can conclude that
-it must be valid.
+counter examples to the original formula and, therefore, it must be valid.
 
-\section{}
+\subsection{GHC's Constraint Solver}
 
-  * Implication constraints
-  * Consistency of assumptions
-  * Consistency
+In this Section, we present a brief overview of some of the functionality
+in GHC's constraint solver.  The full details \cite{outsidein} are beyond
+the scope of this paper.
 
-  * Relation to Selson Oppen
+\paragraph{Implication Constraints.}
+During type inference, GHC uses {\em implication constraints}, which do
+not appear in Haskell source code directly. An implication constraint is,
+roughly, of the form $G\implies W$, where $W$ is a collection
+of constraints that need to be discharged, and $G$ are assumptions that
+may be used while discharging $W$.  In the GHC source code, the constraints
+in $G$ are referred to as {\em given constraints}, while the ones in $W$ are
+known as {\em wanted constraints}.  The intuition behind an implication
+constraint is that $W$ contains the constraints that were collected
+while checking a program fragment, while the constraints in $G$ are local
+assumptions that are only available in this particular piece of code.
+For example, if we are checking the body of a case branch that matches
+on a GADT constructor, then $G$ would be assumptions that we
+obtain from the constructor, while $W$ would be constraints collected
+while analyzing the body of the branch.
+
+\paragraph{The Constraint Solver Pipeline.}
+The constraint solver has two central pieces of state: the {\em work queue},
+and the {\em inert set}.  The work queue contains constraints that need to be
+processed, while the inert set contains constraints that have already
+been processed.  Constraints are removed---one at a time---from the work queue,
+and {\em interacted} with the solver's state.  In the process of interaction
+we may solve a wanted constraint, possibly adding new constraints to the
+work queue, or notice that a constraint is impossible to solve and record
+an error, or---if nothing interesting happens---we add the constraint
+to the inert set.  It is also possible that during interaction a previously
+inert constraint may be reactivated in an "improved" form and re-inserted
+in the work queue.  A single invocation of the constraint solver keeps
+interacting constraints until the work queue is empty and all constraints
+are in the inert set.
+
+\paragraph{Derived Constraints.}
+
+\section{Type Checker Plug-Ins}
 
 \bibliographystyle{abbrvnat}
+\bibliography{refs}
 
-% The bibliography should be embedded for final submission.
-
-\begin{thebibliography}{}
-\softraggedright
-
-\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-P. Q. Smith, and X. Y. Jones. ...reference text...
-
-\end{thebibliography}
+% 
+% % The bibliography should be embedded for final submission.
+% 
+% \begin{thebibliography}{}
+% \softraggedright
+% 
+% \bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
+% P. Q. Smith, and X. Y. Jones. ...reference text...
+% 
+% \end{thebibliography}
 

### 2015-03-10 tests/should-compile/I2.hs
Add tests for things that appear to have become broken.

@@ -1,0 +1,17 @@
+{-# OPTIONS_GHC -fplugin=TypeNatSolver #-}
+{-# LANGUAGE DataKinds #-}
+{-# LANGUAGE TypeOperators #-}
+{-# LANGUAGE TypeFamilies #-}
+{-# LANGUAGE GADTs #-}
+{-# LANGUAGE KindSignatures #-}
+module I2 where
+
+import GHC.TypeLits
+import Data.Proxy
+
+f :: Proxy (a + 6) -> ()
+f _ = ()
+
+g :: (6 <= x) => Proxy x -> ()
+g x = f x
+

### 2015-03-10 tests/should-fail/I1.hs

@@ -1,0 +1,20 @@
+{-# OPTIONS_GHC -fplugin=TypeNatSolver #-}
+{-# LANGUAGE DataKinds #-}
+{-# LANGUAGE TypeOperators #-}
+{-# LANGUAGE TypeFamilies #-}
+{-# LANGUAGE GADTs #-}
+{-# LANGUAGE KindSignatures #-}
+module I1 where
+
+import GHC.TypeLits
+import Data.Proxy
+
+data Nat1 :: Nat -> * where
+  Zero :: Nat1 0
+  Succ :: Nat1 n -> Nat1 (n + 1)
+
+add :: Nat1 m -> Nat1 n -> Nat1 (m + n + 1)
+add Zero x      = x
+add (Succ x) y  = Succ (add x y)
+
+

### 2015-03-11 src/TypeNatSolver.hs
Workaround a bug in GHC.
Fixes #1

@@ -16,6 +16,7 @@ import TcRnMonad ( TcPlugin(..), TcPluginResult(..)
                  , Ct(..), CtEvidence(..), CtLoc, ctLoc, ctPred
                  , mkNonCanonical, isTouchableTcM, unsafeTcPluginTcM
                  )
+import TcRnTypes ( isDerivedCt )
 import Plugins    ( CommandLineOption, defaultPlugin, Plugin(..) )
 
 import TcTypeNats ( typeNatAddTyCon
@@ -37,7 +38,7 @@ import           Data.Map ( Map )
 import qualified Data.Map as Map
 import           Data.IORef ( IORef, newIORef, readIORef, writeIORef
                             , modifyIORef', atomicModifyIORef' )
-import           Data.List ( find )
+import           Data.List ( find, partition )
 import           Data.Maybe ( isNothing )
 import           Data.Either ( partitionEithers )
 import           Control.Monad(ap, liftM, zipWithM)
@@ -84,7 +85,7 @@ pluginStop s = do _ <- tcPluginIO (SMT.stop (solver s))
                   return ()
 
 pluginSolve :: S -> [Ct] -> [Ct] -> [Ct] -> TcPluginM TcPluginResult
-pluginSolve s gs ds ws =
+pluginSolve s gs ds0 ws0 =
   solverDebugFun s "pluginSolve" $
   do resetImportS s
      dbg $ text "-- Givens ------------------------"
@@ -108,6 +109,10 @@ pluginSolve s gs ds ws =
      return res
 
   where
+  -- XXX: This is a temporary hack, because there appears to be a bug in GHC.
+  (ds1,ws) = partition isDerivedCt ws0
+  ds       = ds1 ++ ds0
+
   dbg = tcPluginTrace "NAT"
 
   ppCts [] = text "(empty)"

### 2015-03-11 docs/paper.tex
Checkpoint

@@
 arithmetic.  The technique, however, is more general and makes it
-possible to experiment with other decision procedures supported by
-the SMT solver.
+possible to experiment with other external decision procedures.
 \end{abstract}
@@
 
-\section{Background}
+XXX: Structure of the paper
 
-\subsection {SMT Solvers}
+\section {SMT Solvers}
+
+This Section contains an introduction to the core functionality
+of a typical SMT solver, and may be skipped by readers who are already
+familiar with similar tools.
 
@@
 \cite{z3}, implement a wide collection of decision procedures that
-work together to solve a common problem.  They have proved a useful
-tool in both software and hardware verification.  In this Section,
-we present a very brief introduction to the core functionality
-of a typical SMT solver---readers familiar with the tools could
-skip this Section.
-
+work together to solve a common problem.  They have proved to be a useful
+tool in both software and hardware verification.
 From a user's perspective, the core functionality of an SMT solver is fairly
@@
 
-The example declares a constant, $x$, asserts some formulas
+The example declares an integer constant, $x$, asserts some formulas---%
+using a prefix notation---%
 about it, and then asks the solver if the asserted formulas are
@@
 the uninterpreted constants to concrete values that make the
-asserted formulas true.
+formulas true.
 
-It is common to use the same machinery for proving the validity of
-universally quantified formulas by looking for counter-examples.
+The same machinery may also be used to prove the validity of
+a universally quantified formula. The idea is that we use the SMT solver
+to look for a {\em counter-example} to the formula, and if on such
+example exists, then we can conclude the formula is valid.
 For example, if we want to prove that $\forall x. (3 + x = 8) \implies x = 5$,
-then we can use the SMT solver to try to find some $x$ that contradicts
-the formula:
+then we can use the SMT solver to try to find some $x$ that contradicts it:
 
@@
 To invalidate an implication, we need to assume the premise, and try to
-invalidate the conclusion, which is why the second assertion in the
-example is negated. In this case the SMT solver will tell us that the
-asserted formulas are not satisfiable, which means that there are no
-counter examples to the original formula and, therefore, it must be valid.
+invalidate the conclusion, which is why the second assertion is negated.
+In this case the SMT solver tells us that the asserted formulas are not
+satisfiable, which means that there are no counter examples to the original
+formula and, therefore, it must be valid.
 
-\subsection{GHC's Constraint Solver}
+\section{GHC's Constraint Solver}
 
-In this Section, we present a brief overview of some of the functionality
-in GHC's constraint solver.  The full details \cite{outsidein} are beyond
-the scope of this paper.
+In this Section, we present relevant aspects of GHC's constraint solver.
+The full details of the algorithm \cite{outsidein} are beyond the scope
+of this paper.
 
-\paragraph{Implication Constraints.}
+\subsection{Implication Constraints}
 During type inference, GHC uses {\em implication constraints}, which do
@@
 constraint is that $W$ contains the constraints that were collected
-while checking a program fragment, while the constraints in $G$ are local
-assumptions that are only available in this particular piece of code.
-For example, if we are checking the body of a case branch that matches
-on a GADT constructor, then $G$ would be assumptions that we
-obtain from the constructor, while $W$ would be constraints collected
-while analyzing the body of the branch.
+while checking a program fragment, while $G$ contains local assumptions
+that are available only in this particular piece of code.  For example,
+consider the following program fragment, which uses a GADT:
+\begin{Verbatim}
+data E :: * -> * where
+  EInt :: Int -> E Int
 
-\paragraph{The Constraint Solver Pipeline.}
+isZero :: E a -> Bool
+isZero (EInt x) = x == 0
+\end{Verbatim}
+When we check the definition of \verb"isZero", we end up with an implication
+constraint like this:
+\begin{Verbatim}
+(a ~ Int) => (Num a, Eq a)
+\end{Verbatim}
+Here, \verb"a" is the type of the pattern variable \verb"x", the wanted
+constraints arise from the use of \verb"0" and \verb"(==)" respectively,
+and the given constraint is obtain by pattern matching with \verb"EInt".
+
+\subsection{The Constraint Solver State}
+Implication constraints are solved by two calls to the constraint solver:
+the first call processes (i.e., assumes) the given constraints, and the
+second one processes the wanted constraints.
+
 The constraint solver has two central pieces of state: the {\em work queue},
@@
 to the inert set.  It is also possible that during interaction a previously
-inert constraint may be reactivated in an "improved" form and re-inserted
+inert constraint may be reactivated in a rewritten form and re-inserted
 in the work queue.  A single invocation of the constraint solver keeps
@@
 
-\paragraph{Derived Constraints.}
+\subsection{Type-Checker Plugins}
+Integrating an SMT solver with GHC's type-checker is not the only extension
+one might want to have---indeed, there is existing work that extends GHC
+with support for units of measure \cite{units-of-measure}.  Instead of
+having many ad-hoc extensions directly in GHC's constraint solver, we
+collaborated to define and implement and API for extending GHC's functionality
+via {\em type-checker plug-ins}.  At present, we are aware of two main
+users of this API---our work, and the work on units of measure, but we
+hope that this mechanism would make it easier for researcher to experiment
+with various extensions to GHC's constraint solver.
+
+An interesting question about type-checker plug-ins is: at what point in
+GHC's type checker should we hook them in? We considered a few alternatives:
+\begin{enumerate}
+\item Add a pass to the solver's pipeline, meaning that plug-ins have to
+work with constraints one at a time.
+\item Add a call to the plug-ins after the constraints solver has reached
+an inert state.
+\item Hand off implication constraints directly to the plug-ins,
+before invoking GHC's constraint solver.
+\end{enumerate}
+
+At the end, we decided on option 2.  Option 3 is the most general, as it
+would allow for the plug-in to completely override GHC's behavior.  However,
+we were more interested in {\em extending} GHC's capabilities rather than
+{\em replacing} them, so we opted against it.  Option 1, on the other hand,
+has the closest integration with GHC, but since plug-ins need to precess
+constraints one at a time, then they often ended up needing some state,
+which than had to be stored and managed somewhere, which became rather
+complicated.
+
+We chose option 2 as a nice middle ground, which allows us to look at
+all constraints at once, but plug-ins can assume that standard stuff
+the GHC knows about has already been eliminated. So, after the constraint
+solver reaches an inert state, it calls into the plug-ins, which examine
+the inert state and attempt to make progress.  If they do, then the
+constraint solver is restarted and process repeats.
+
+\subsection{Improvement and Derived Constraints}
+An improving substitution \cite{improvement} allows us to instantiate variables
+in the constraints under consideration with types, thus potentially enabling
+further constraint simplification.  Of course, we should only do so, as long
+as we preserve soundness and completeness.  In this context preserving
+soundness means that the new constraints should imply the original
+constraints (i.e., we didn't just drop some constraints),
+while completeness means that the original constraints imply the new ones
+(i.e., we didn't just make some arbitrary assumptions).
+
+In GHC, improvement happens by using {\em equality constraints} to rewrite
+constraints.  There are three sources of equality constraints:
+\begin{itemize}
+\item given equalities are computed from the given constraints
+\item wanted equalities arise
+\item derived equalities
+\end{itemize}
+
+\subsection{Improvement}
 
 \section{Type Checker Plug-Ins}
+
+\section{Algorithm}
+
+\subsection{Idnetifying Constraints that Belong to the Theory}
+
+\subsection{Consistency and Improvement}
+
+\subsection{Simplifying Constraints}
+
+Given an implication constraint: $G\implies W$.
+\begin{enumerate}
+\item Identify constraints in the theory
+\item Name sub-terms that are outside the theory
+\item Assume $G$
+\item Check for satisfiability
+  \begin{itemize}
+  \item If unsat, implication is trivial, record error
+  \item If sat, use model for improvement
+  \end{itemize}
+\item ...
+\end{enumerate}
+
+\section{A Modular Type Checker}
+
+Relation between type-checker and
+the Nelson-Oppen method for combining
+decision procedures.
+
+\section {Conclusions}
 

### 2015-03-12 docs/paper.tex
Checkpoint

@@
 \section{GHC's Constraint Solver}
+\label{GHC}
 
@@
 \end{Verbatim}
-When we check the definition of \verb"isZero", we end up with an implication
+When we check the definition of \Verb"isZero", we end up with an implication
 constraint like this:
@@
 \end{Verbatim}
-Here, \verb"a" is the type of the pattern variable \verb"x", the wanted
-constraints arise from the use of \verb"0" and \verb"(==)" respectively,
-and the given constraint is obtain by pattern matching with \verb"EInt".
+Here, \Verb"a" is the type of the pattern variable \Verb"x", the wanted
+constraints arise from the use of \Verb"0" and \Verb"(==)" respectively,
+and the given constraint is obtain by pattern matching with \Verb"EInt".
 
@@
 An improving substitution \cite{improvement} allows us to instantiate variables
-in the constraints under consideration with types, thus potentially enabling
-further constraint simplification.  Of course, we should only do so, as long
+in the constraints with types, potentially enabling further constraint
+simplification.  Of course, we should only do so, as long
 as we preserve soundness and completeness.  In this context preserving
@@
 while completeness means that the original constraints imply the new ones
-(i.e., we didn't just make some arbitrary assumptions).
+(i.e., we didn't loose generality by making arbitrary assumptions).
 
-In GHC, improvement happens by using {\em equality constraints} to rewrite
-constraints.  There are three sources of equality constraints:
+In GHC, improvement happens when we rewrite constraints using
+{\em equality constraints}. There are three sources of equality constraints:
 \begin{itemize}
-\item given equalities are computed from the given constraints
-\item wanted equalities arise
-\item derived equalities
+\item {\em given equalities} are implied by the given constraints,
+\item {\em wanted equalities} arise when solving wanted constraints,
+\item {\em derived equalities} are implied by the given and wanted constraints,
+together.
 \end{itemize}
 
-\subsection{Improvement}
+The provenance of an equality constraint determines the kinds of
+improvements that we can use it for.  Given equalities have solid proof,
+and so we may use congruence to rewrite any other constraint.
+On the other hand, wanted constraints are goals that need to be proved,
+so they {\em cannot} be used to rewrite given constraints.  We may still
+use them to rewrite other wanted or derived constraints though.
+Finally, derived equalities are implied by the given and wanted constraints
+jointly, so they may not be used to rewrite given or wanted constraints
+directly, as doing so may lead to circular reasoning.
 
-\section{Type Checker Plug-Ins}
+Instead, derived equalities help with type inference, by guiding the
+instantiation of {\em unification variables} with types. In general,
+it is always sound to instantiate a unification variable with whatever
+type we want: doing so may reject valid programs, but it will not accept
+invalid ones, because we still need to solve all necessary constraints.
+Of course, we don't want to reject valid programs, and this is where derived
+equalities help us: since they are implied by the goals and assumptions
+together, we know that we are not loosing any generality when instantiating
+as suggested by a derived constraint.  So, for example, if we compute a
+derived constraint \Verb"x ~ Int", and we have a wanted constraint
+\Verb"Eq x", and \Verb"x" is a unification variable, then we may
+rewrite \Verb"Eq x" to \Verb"Eq Int", which we would proceed to solve
+as usual.
 
-\section{Algorithm}
+\section{Integrating GHC with an SMT Solver}
 
-\subsection{Idnetifying Constraints that Belong to the Theory}
+We describe the algorithm in the context of the theory of linear arithmetic
+over natural numbers, as having a concrete theory makes it easier to explain
+the process and illustrate it with examples.  In the next Section,
+we discuss how, and why, we might want to consider other theories also.
+
+\paragraph{Input.} Currently, the algorithm is implemented as a type-checker
+plug-in, and so the input to the algorithm is a collection of constraints
+that GHC has determine to be inert: GHC simplified everything as much
+as it could, improved using equalities, and basically cannot see anything
+else to do.  The inert constraints are presented to the plug-in in three
+collections, based on the constraints' provenance: one group contains
+the given constraints (i.e., assumption), one group contains derived
+constraints (see Section~\ref{GHC} for details), and one group contains
+the wanted constraints, which are the goals that need solving.
+
+\paragraph{Output.} The desired output of the algorithm is as follows:
+\begin{itemize}
+  \item Solve as many wanted constraints as possible.
+  \item Notice if the wanted constraints are inconsistent.
+  \item Compute new given and derived equalities to help solve constraints
+        that are outside this decision procedure's scope.
+\end{itemize}
+The first bullet is clearly the main purpose of the algorithm, but the other
+two are quite important also.
+
+The second bullet is useful because it helps us avoid inferring type
+schemes with unsatisfiable constraints.  Suppose, for example, that
+we had just a single wanted constraint: \Verb"(x + 5) ~ 2".  Since we
+are working with natural numbers, this constraint has no solution, so
+we will not be able to solve it. As a result, we may end up inferring
+a type like this:
+\begin{Verbatim}
+someFun :: forall x. (x + 5) ~ 2 => ...
+\end{Verbatim}
+While, technically, this is not wrong, it is undesirable because the type
+error is clearly in the definition of \Verb"someFun", but we would delay
+reporting the error until the function is called.  So, it is better if
+we noticed that the wanted constraints are not satisfiable, are report
+and error immediately.
+
+The importance of the third bullet is that it enables collaboration between
+the SMT solver plug-in, and the rest of GHC (i.e., the main constraint solver,
+and also other plug-ins).  For example, consider an implication constraint
+of the form:
+\begin{Verbatim}
+forall x. (x + 5) ~ 8  =>  KnownNat x
+\end{Verbatim}
+In this case, we'd have to use the SMT solver to compute a new given
+equality, \Verb"x ~ 3".  This equality will then be used by GHC to
+rewrite the wanted constraint \Verb"KnownNat x" to \Verb"KnownNat 3",
+which in turn can be discharged by the custom solver for the \Verb"KnownNat"
+class.  This ``collaboration'' between different solvers is quite common
+in GHC.
+We discuss this observation further in Section~\ref{modular-typechecker}.
+
+\subsection{Identifying Relevant Constraints}
+Our first task is to identify constraints that are relevant to our solver.
+In the current implementation, we consider equalities and inequalities
+between a subset of the types of kind \Verb"Nat".   The kind \Verb"Nat" is
+inhabited by an infinite family of type-constants:
+\begin{Verbatim}
+0, 1, 2, .. :: Nat
+\end{Verbatim}
+These constants may be combined and compared using the following
+type-level functions:
+\begin{Verbatim}
+type family (+)   :: Nat -> Nat -> Nat
+type family (*)   :: Nat -> Nat -> Nat
+type family (<=?) :: Nat -> Nat -> Bool
+\end{Verbatim}
+These functions have no user-specified definitions: instead, we extended
+the core GHC simplifier with support for forward evaluation on concrete
+values, so it will evaluate concrete expressions, such as \Verb"2 + 3".
+This simple forward evaluation simplifies and cleans up the constraints,
+leaving the more complex reasoning involving variables to the algorithm
+being presently described.
+
+The declaration for \Verb"(<=?)" refers to the kind \Verb"Bool", which
+is simply the lifted \Verb"Bool" type.  As expected, it is inhabited by the
+empty types \Verb"True" and \Verb"False".
+
+So the language of interest to us it constraints formed with the two
+predicate symbols, \Verb"(~)" and \Verb"(<=?)", and types made out
+of variables, natural number constants, and the operations \Verb"(+)" and
+\Verb"(*)".  Furthermore, since we are restricted to working with {\em linear}
+arithmetic, we only consider multiplication by a constant.
+
+Of course, constraints may contain other symbols too.  For example,
+programmers are free to define their own type functions that return
+results of kind \Verb"Nat", so it is entirely possible to encounter
+constraints like \Verb"(F 3 + 4) ~ 7", where \Verb"F" is some programmer-defined
+type-family.
+
+The process of identifying relevant constraints proceeds as follows:
+\begin{enumerate}
+\item Eliminate constraints that do not have a relevant top-level predicate
+symbol.  The relevant predicate symbols are \Verb"(<=?)", and \Verb"(~)" when
+used at kind \Verb"Nat".
+\item Import the parameters to the predicates guided by the kind:
+  \begin{itemize}
+  \item The two supported kinds are \Verb"Nat" and \Verb"Bool"
+  \item Currently, there are no interesting operations at kind \Verb"Bool"
+  \item If we encounter a type outside of our theory, we name it, and replace
+        it with a temporary variable.
+  \item We remember the name assignments, so that if the same ``foreign'' type
+        appears multiple times, it will get the same name.
+  \end{itemize}
+\end{enumerate}
+
+The step of naming foreign types is similar to the flattening step performed
+by GHC's constraint solver, the main difference being that GHC names every
+call to a type-function, while we name only terms outside of the theory.
+Still, this duplication of work is somewhat unsatisfactory, and it would
+be nice to extend GHC's flattening pass so that it can be reused by
+external plug-ins.
 
@@
 
-\section{A Modular Type Checker}
+\section{Other Theories}
+
+\paragraph{Integers.}
+\paragraph{Boolean.}  SAT solver in the type system
+\paragraph{Bit-vectors.}  Aribtrary modulo arithmeitc; finite sets of effects.
+\paragraph{Sets.}
+
+\section{A Modular Constraint Solver}
+\label{modular-typechecker}
 

### 2015-03-12 docs/paper.tex
A bit about incremental solvers

@@
 
+\subsection{The Core Functionality}
+
 SMT solvers, such as CVC4 \cite{cvc4}, Yices \cite{yices}, and Z3
@@
 (declare-fun x () Int)
-(assert (= 3 x) 8)
+(assert (= (+ 3 x) 8))
 (assert (not (= x 5)))
@@
 formula and, therefore, it must be valid.
+
+\subsection{Incremental Solvers}
+
+Many SMT solvers have support for asserting formulas {\em incrementally}.
+This means that the solver performs a little work every time a new formula
+is asserted, rather than collecting all formulas and doing all the work
+in batch mode, once we ask about satisfiability.  Furthermore, incremental
+solvers have the ability to mark a particular solver state, and then revert
+back to it, using a stack discipline.  Here is an example:
+
+\begin{Verbatim}
+(declare-fun x () Int)
+(assert (= (+ 5 x) 3))
+(push 1)
+  (assert (>= x 0))
+  (check-sat)
+(pop 1)
+(check-sat)
+\end{Verbatim}
+
+This example asks the solver two questions, and the answer to the first
+one is \Verb"unsatisfiable", while the answer to the second one is
+\Verb"satisfiable".
+
+The \Verb"push" command instructs to solver to save its state, then
+proceed as normal.  When the solver encounters a \Verb"pop" command,
+it reverts to the last saved state.
+
+This functionality is extremely useful when we want to ask many questions
+that are largely the same, and differ only in a few assertions:
+we can perform the majority of the work once, and then use
+\Verb"push" and \Verb"pop" to just assert the differences.
+In our small example, the work for the first assertion shared in
+both calls to \Verb"check-sat".
 
@@
 we noticed that the wanted constraints are not satisfiable, are report
-and error immediately.
+an error immediately.
 
 The importance of the third bullet is that it enables collaboration between
-the SMT solver plug-in, and the rest of GHC (i.e., the main constraint solver,
-and also other plug-ins).  For example, consider an implication constraint
+the plug-in and the rest of GHC (i.e., the main constraint solver,
+and other plug-ins).  For example, consider an implication constraint
 of the form:
@@
 which in turn can be discharged by the custom solver for the \Verb"KnownNat"
-class.  This ``collaboration'' between different solvers is quite common
-in GHC.
-We discuss this observation further in Section~\ref{modular-typechecker}.
+class.  Such collaboration between different solvers is quite common.
+Interestingly, this looks a lot like the collaboration between decision
+procedures in an SMT solver!  We discuss this observation further
+in Section~\ref{modular-typechecker}.
 

### 2015-03-13 docs/paper.tex
More text

@@
 \usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{amsthm}
 \usepackage{fancyvrb}
 \usepackage{url}
+
+\newtheorem*{lemma}{Lemma}
+\newtheorem*{corollary}{Corollary}
 
@@
 We present a technique for integrating GHC's type-checker with the
-functionality provided by an SMT solver.  The technique was developed
-to add support for reasoning about type-level functions on natural
-numbers, and so our implementation mainly uses the theory of linear
-arithmetic.  The technique, however, is more general and makes it
-possible to experiment with other external decision procedures.
+an SMT solver.  The technique was developed to add support for reasoning
+about type-level functions on natural numbers, and so our implementation
+uses the theory of linear arithmetic.  The technique, however, is more
+general and makes it possible to experiment with other external decision
+procedures, such as reasoning about type-level booleans, bit-vectors,
+or any other theory supported by SMT solvers.
 \end{abstract}
@@
 \section {SMT Solvers}
+\label{smt}
 
@@
 one is \Verb"unsatisfiable", while the answer to the second one is
-\Verb"satisfiable".
+\Verb"satisfiable" and the solution is \Verb"x = -2".
 
-The \Verb"push" command instructs to solver to save its state, then
+The \Verb"push" command instructs the solver to save its state, then
 proceed as normal.  When the solver encounters a \Verb"pop" command,
@@
 \Verb"push" and \Verb"pop" to just assert the differences.
-In our small example, the work for the first assertion shared in
+In our small example, the work for the first assertion is shared in
 both calls to \Verb"check-sat".
@@
 and {\em interacted} with the solver's state.  In the process of interaction
-we may solve a wanted constraint, possibly adding new constraints to the
-work queue, or notice that a constraint is impossible to solve and record
-an error, or---if nothing interesting happens---we add the constraint
-to the inert set.  It is also possible that during interaction a previously
+we may solve constraint, generate new work, or report impossible constraints.
+If nothing interesting happens, then we relocate the constraint to the
+inert set.  It is also possible that during interaction a previously
 inert constraint may be reactivated in a rewritten form and re-inserted
@@
 interacting constraints until the work queue is empty and all constraints
-are in the inert set.
+become inert.
 
 \subsection{Type-Checker Plugins}
-Integrating an SMT solver with GHC's type-checker is not the only extension
-one might want to have---indeed, there is existing work that extends GHC
-with support for units of measure \cite{units-of-measure}.  Instead of
-having many ad-hoc extensions directly in GHC's constraint solver, we
-collaborated to define and implement and API for extending GHC's functionality
-via {\em type-checker plug-ins}.  At present, we are aware of two main
-users of this API---our work, and the work on units of measure, but we
-hope that this mechanism would make it easier for researcher to experiment
-with various extensions to GHC's constraint solver.
+In this paper we describing a fairly general extension to the constraint
+solver, however one could always imagine others too. Indeed, there is
+existing work that extends GHC with support for units of measure
+\cite{units-of-measure}.  Instead of having many ad-hoc extensions directly
+in GHC's constraint solver, we collaborated to define and implement an
+API for extending GHC's functionality via {\em type-checker plug-ins}.
+At present, we are aware of two main
+users of this API---our work, and the work on units of measure. We
+hope, however, that this mechanism would be useful to other researchers too,
+as it makes it fairly easy to experiment with various extensions to
+GHC's constraint solver.
 
 An interesting question about type-checker plug-ins is: at what point in
-GHC's type checker should we hook them in? We considered a few alternatives:
+GHC's type checker should we invoke them?  The answer to this question
+affects their functionality, and the interface that a plug-in would have
+to implement.  We considered a few alternatives:
 \begin{enumerate}
-\item Add a pass to the solver's pipeline, meaning that plug-ins have to
-work with constraints one at a time.
+\item Add a pass to the solver's pipeline, meaning that plug-ins process
+constraints one at a time, the way GHC does.
 \item Add a call to the plug-ins after the constraints solver has reached
@@
 
-At the end, we decided on option 2.  Option 3 is the most general, as it
+Option 1 has the closest integration with GHC, but since plug-ins need
+to process constraints one at a time, then they often ended up needing
+some state, which than has to be stored and managed somewhere,
+which became rather complicated.
+
+Option 3 is the most general, as it
 would allow for the plug-in to completely override GHC's behavior.  However,
 we were more interested in {\em extending} GHC's capabilities rather than
-{\em replacing} them, so we opted against it.  Option 1, on the other hand,
-has the closest integration with GHC, but since plug-ins need to precess
-constraints one at a time, then they often ended up needing some state,
-which than had to be stored and managed somewhere, which became rather
-complicated.
+{\em replacing} them, so we opted against it.
 
-We chose option 2 as a nice middle ground, which allows us to look at
-all constraints at once, but plug-ins can assume that standard stuff
-the GHC knows about has already been eliminated. So, after the constraint
+We chose option 2 as a nice middle ground. It allows a plug-in to
+work with all constraints at once, but it has the benefit that standard work
+done by GHC has already happened. So, after the constraint
 solver reaches an inert state, it calls into the plug-ins, which examine
@@
 simplification.  Of course, we should only do so, as long
-as we preserve soundness and completeness.  In this context preserving
-soundness means that the new constraints should imply the original
+as we preserve soundness and completeness.  In this context, preserving
+soundness means that the improved constraints should imply the original
 constraints (i.e., we didn't just drop some constraints),
-while completeness means that the original constraints imply the new ones
+while completeness means that the original constraints imply the improved ones
 (i.e., we didn't loose generality by making arbitrary assumptions).
 
-In GHC, improvement happens when we rewrite constraints using
-{\em equality constraints}. There are three sources of equality constraints:
+In GHC, improvement happens by rewriting with {\em equality constraints}.
+There are three sources of equality constraints:
 \begin{itemize}
@@
 We describe the algorithm in the context of the theory of linear arithmetic
-over natural numbers, as having a concrete theory makes it easier to explain
+over the natural numbers, as having a concrete theory makes it easier to explain
 the process and illustrate it with examples.  In the next Section,
@@
 plug-in, and so the input to the algorithm is a collection of constraints
-that GHC has determine to be inert: GHC simplified everything as much
-as it could, improved using equalities, and basically cannot see anything
-else to do.  The inert constraints are presented to the plug-in in three
-collections, based on the constraints' provenance: one group contains
-the given constraints (i.e., assumption), one group contains derived
+that GHC has determine to be inert. This means that GHC simplified everything
+as much as it could, improved using equalities, and basically cannot see
+anything else to do.  The inert constraints are presented to the plug-in
+in three collections, based on the constraints' provenance: one group contains
+the given constraints (i.e., assumptions), one group contains derived
 constraints (see Section~\ref{GHC} for details), and one group contains
@@
 \end{itemize}
-The first bullet is clearly the main purpose of the algorithm, but the other
+The first task is the most obvious purpose of the algorithm, but the other
 two are quite important also.
 
-The second bullet is useful because it helps us avoid inferring type
-schemes with unsatisfiable constraints.  Suppose, for example, that
-we had just a single wanted constraint: \Verb"(x + 5) ~ 2".  Since we
-are working with natural numbers, this constraint has no solution, so
-we will not be able to solve it. As a result, we may end up inferring
-a type like this:
+Noticing inconsistencies avoids the inference of types with unsatisfiable
+constraints.  Consider, for example, a single wanted constraint,
+\Verb"(x + 5) ~ 2".  Since we are working with natural numbers,
+this constraint has no solution, so we cannot solve it.
+As a result, we may end up inferring a type like this:
 \begin{Verbatim}
@@
 error is clearly in the definition of \Verb"someFun", but we would delay
-reporting the error until the function is called.  So, it is better if
-we noticed that the wanted constraints are not satisfiable, are report
-an error immediately.
+reporting the error until the function is called.  So, we'd like to
+notice constraints that are impossible to solve (i.e., they are logically
+equivalent to $\bot$), so that we can report type-errors that are closer
+to their true location.  Note that the source of a contradiction may
+be a combination of constraints, and not just a single one.  For example,
+the constraints \Verb"x >= 5" and \Verb"3 >= x" are inconsistent together,
+but have solutions when considered individually.
 
-The importance of the third bullet is that it enables collaboration between
-the plug-in and the rest of GHC (i.e., the main constraint solver,
+Computing new equations is also very important, as it enables collaboration
+between the plug-in and the rest of GHC (i.e., the main constraint solver,
 and other plug-ins).  For example, consider an implication constraint
@@
 \end{Verbatim}
-In this case, we'd have to use the SMT solver to compute a new given
-equality, \Verb"x ~ 3".  This equality will then be used by GHC to
-rewrite the wanted constraint \Verb"KnownNat x" to \Verb"KnownNat 3",
-which in turn can be discharged by the custom solver for the \Verb"KnownNat"
+In this case, we'd like to use the SMT solver to compute a new given
+equality, \Verb"x ~ 3".  Then, this equality can be used by GHC to
+rewrite the wanted constraint \Verb"KnownNat x" to \Verb"KnownNat 3"
+which, in turn, can be discharged by the custom solver for the \Verb"KnownNat"
 class.  Such collaboration between different solvers is quite common.
@@
 
-\subsection{Identifying Relevant Constraints}
+% \subsection{Communicating With The Solver}
+% Our implementation uses the \Verb"simple-smt" library \cite{simple-smt}
+% to communicate with external SMT solvers.
+
+\subsection{The Language of Constraints}
 Our first task is to identify constraints that are relevant to our solver.
@@
 values, so it will evaluate concrete expressions, such as \Verb"2 + 3".
-This simple forward evaluation simplifies and cleans up the constraints,
-leaving the more complex reasoning involving variables to the algorithm
-being presently described.
+This simple forward evaluation cleans up the constraints, leaving the more
+complex reasoning---involving variables---to the algorithm being presently
+described.
 
@@
 is simply the lifted \Verb"Bool" type.  As expected, it is inhabited by the
-empty types \Verb"True" and \Verb"False".
+empty types \Verb"True" and \Verb"False".   Having \Verb"(<=?)" be a function
+that returns a boolean is a little more convenient for programmers than having
+a constraint. In this form, a programmer may inspect the result of the
+inequality and make a decision based on it.  We can use a simple
+abbreviation to implement the corresponding inequality constraint:
+\begin{Verbatim}
+type (<=) a b  = (a <=? b) ~ True
+\end{Verbatim}
 
-So the language of interest to us it constraints formed with the two
-predicate symbols, \Verb"(~)" and \Verb"(<=?)", and types made out
-of variables, natural number constants, and the operations \Verb"(+)" and
-\Verb"(*)".  Furthermore, since we are restricted to working with {\em linear}
-arithmetic, we only consider multiplication by a constant.
+The constraints of interest to the algorithm are summarized by the following
+grammar:
+\begin{align*}
+c & = e \overset {\mathbb{N}} \sim e ~|~  e \overset {\mathbb{B}} \sim e \\
+e^\mathbb{N} & = \alpha ~|~ n ~|~ e + e ~|~ n * e \\
+e^\mathbb{B} & = \alpha ~|~ \mathrm{False} ~|~ \mathrm{True} ~|~ e \leq^? ~e \\
+n          & = 0 ~|~ 1 ~|~ \dots
+\end{align*}
 
-Of course, constraints may contain other symbols too.  For example,
+\subsection{Importing Constraints}
+
+Of course, general constraints may be more complex.  In particular,
 programmers are free to define their own type functions that return
@@
 constraints like \Verb"(F 3 + 4) ~ 7", where \Verb"F" is some programmer-defined
-type-family.
+type-family.  It is also possible to encounter non-linear constraints such
+as \Verb"x * y".
 
-The process of identifying relevant constraints proceeds as follows:
+The process of identifying relevant constraints is as follows:
 \begin{enumerate}
-\item Eliminate constraints that do not have a relevant top-level predicate
-symbol.  The relevant predicate symbols are \Verb"(<=?)", and \Verb"(~)" when
-used at kind \Verb"Nat".
-\item Import the parameters to the predicates guided by the kind:
-  \begin{itemize}
-  \item The two supported kinds are \Verb"Nat" and \Verb"Bool"
-  \item Currently, there are no interesting operations at kind \Verb"Bool"
-  \item If we encounter a type outside of our theory, we name it, and replace
-        it with a temporary variable.
-  \item We remember the name assignments, so that if the same ``foreign'' type
-        appears multiple times, it will get the same name.
-  \end{itemize}
+\item Set aside constraints that do not have a relevant top-level predicate
+symbol.
+\item Import the parameters to the predicates guided by the kind.
+\item If we encounter a type outside of our theory, then we name it and replace
+      it with a variable.
+\item We remember the names assigned to various types, so that we may reuse
+      the same name if a type appears multiple times.
 \end{enumerate}
 
+For example, the constraint \Verb"(F 3 + 4) ~ 7" will be imported as
+\Verb"(x + 4) ~ 7", and we will remember that \Verb"x" stands for \Verb"F 3".
+Later on, if we need to return results to the core GHC solver that mention
+\Verb"x", we replace it by the orignal expression, \Verb"F 3".
+
+The process of abstracting foreign constraints makes sense in our context,
+because it {\em generalizes} the constraint.  Consider a constraint $C(t)$, and
+its generalized form $C(x)$, where the sub-term $t$ is replaced by
+the variable $x$, which does not occur in $C(t)$.
+
+\begin{lemma}
+If $C(t)$ is satisfiable, then $C(x)$ is also satisfiable.
+\begin{proof} This follows because, by definition,
+$C(t) \iff C(x) \wedge x = t$. Therefore, if $\sigma$ is satisfying
+assignment for $C(t)$, then $\sigma \cup \{ x = \sigma t \}$ is satisfying
+assignment for $C(x)$.
+\end{proof}
+\end{lemma}
+
+\begin{corollary} If $C(x)$ is not satisfiable, then neither is $C(t)$.
+This is simply the contrapositive form of the previous Lemma.
+\end{corollary}
+
+Recall from Section~\ref{smt}, than to prove something, we make sure
+that there are no counter examples.  The Corollary states that if there
+are no counter examples to the abstracted constraint, than there will
+be no counter-examples to the original constraint as well, and so our
+proof is sound.  From a logical stand-point, naming away sub-terms
+amounts to trying to prove a more general fact, and so if we can complete
+the proof in the more general setting, then the proof will work for
+the special case of the original constraint.
+
+\paragraph{Aside.}
 The step of naming foreign types is similar to the flattening step performed
@@
 
-\subsection{Consistency and Improvement}
+\subsection{Checking for Consistency}
 
-\subsection{Simplifying Constraints}
+Given some constraints in our theory, we would like to know if a
+solution exists.  If the constraints are inconsistent, then we can
+terminate the constraint solving problem early, and report an error,
+as discussed previously.
 
-Given an implication constraint: $G\implies W$.
-\begin{enumerate}
-\item Identify constraints in the theory
-\item Name sub-terms that are outside the theory
-\item Assume $G$
-\item Check for satisfiability
-  \begin{itemize}
-  \item If unsat, implication is trivial, record error
-  \item If sat, use model for improvement
-  \end{itemize}
-\item ...
-\end{enumerate}
+To check for consistency, we assert all constraints, and ask the SMT
+solver if the result is satisfiable.  If this is not the case, then
+we report and error, otherwise we proceed with the algorithm.  Note that
+since we are working with generalized constraints, if we find an
+inconsistency, then we are sure that we've detected a problem.  However,
+if the SMT solver says that the constraints are satisfiable, then
+we now that this is the case from the point of view of our theory,
+however the constraints might still be unsatisfiable if one took a global
+view of the problem.  Consider, for example, \Verb"(F a + 4) ~ 7".
+This will get imported into our plug-in as \Verb"(x + 4) ~ 7", which is
+satisfiable by \Verb"x = 3".  Now, some other theory might know that
+\Verb"F a ~ 3" is not actually possible, but we won't detect this problem
+at this stage.
+
+\subsection{Minimizing Conflicts}
+If we detect a contradiction, then
+we know that there is no possible solution to the constraints.  However,
+often only a few of the constraints are involved in the actual problem,
+and it is much nicer if we report only them as the cause of the error,
+rather than getting a huge error involving all constraints.  If an SMT
+solver detects that a collection of assertions is inconsistent, it may
+be able to compute an {\em unsatisfiability core}, which is the sub-set
+of the assertions that lead to the conflict---exactly what we want!
+
+Unfortunately, not all SMT solvers support this feature and CVC4---%
+the SMT solver that we used during the development of this work---%
+does not have this ability.  To work around this, we implemented
+a custom minimization algorithm which work as follows:
+\begin{Verbatim}
+minimize s yes [] = return yes
+minimize s yes mb =
+  do solverPush s
+     search s yes [] mb
+
+search s yes mb (c : cs) =
+  do solverAssert s c
+     res <- solverCheck s
+     case res of
+       Unsat ->
+         do solverPop s
+            solverAssert s c
+            minimize s (c : yes) mb
+
+       _ -> search s yes (c : mb) cs
+\end{Verbatim}
+The algorithm keeps track of constraints that we are certain are part
+of the conflict in the variable \Verb"yes".  In addition, the algorithm
+has the invariant that the constraints in \Verb"yes" are always asserted
+in the solver's state, represented by the variable \Verb"s".
+Initially, \Verb"yes" starts off empty.
+
+We search for conflicts in the function \Verb"search", which asserts
+constraints one at a time, until a conflict is discovered.  The constraint
+that caused the conflict is added to the \Verb"yes" set, and then we
+examine the previously asserted constraint, to check if they are necessary
+for the conflict.
+
+The minimization process performs in $O(n^2)$ time, where $n$ is the
+number of constraints.  This has not been a big problem,
+as this minimization happens only when we report errors and, typically,
+the number of constraints is fairly low.
+
+\subsection{Improvement}
+
+If all constraint are consistent (i.e., have at least one solution),
+then we check for computed equality constraints.  As discussed previously,
+these help other parts of the constraint solver to make progress.
+
+It is convenient to compute the improvements right after the consistency
+check because at this point we already have all constraints asserted
+in the solver's state.  If all constraints in scope are given constraints,
+then we also generate given equalities, otherwise we generate derived
+equalities.
+
+The current implementation considers three forms of improvement,
+two of which are completely generic, and one of which is specific to
+linear arithmetic.
+
+\paragraph{Improve to Constant.} As we know that the constraints are
+consistent, we can ask the SMT solver for a satisfying assignment.
+This assignment contains one possible solution to constraints, but
+we should not emit an improving equality unless we are sure that
+this is the {\em the only} possible solution.  So, if $x = v$ is
+in the satisfying assumption, then we try to prove that the currently
+asserted constraints imply this fact.  We do this by temporarily asserting
+$x \neq v$, and checking for satisfiability.  If the resulting is unsatisfiable,
+then we know that $v$ is the only possible value for $x$,
+and we can emit the corresponding improving equation.
+\begin{Verbatim}
+solverProve s p =
+  do solverPush s
+     solverAssert s (SMT.not p)
+     res <- solverCheck s
+     solverPop s
+     return (res == Unsat)
+
+mustBeK s x v =
+  solverPorve s (eq (SMT.const x) (value v))
+\end{Verbatim}
+This process takes $O(n)$ time, where $n$ is the number of variables.
+Note that here we are making use of the incremental capabilities of the
+solver, and reusing all asserted constraints, just doing one
+additional \Verb"assert" per variable. To see the process in action,
+consider the constraint \Verb"(2 * x) ~ 16".  The steps that we'd perform
+are:
+\begin{Verbatim}
+(assert (= (* 2 x) 16))
+(check-sat)
+; SMT response: SAT
+(get-value (x))
+; SMT response: (8)
+(push 1)
+(assert (not (= x 8)))
+(check-sat)
+; SMT response: UNSAT
+(pop 1)
+; We generate a computed improvement: x ~ 8
+\end{Verbatim}
+
+\paragraph{Improve to Variable.}
+We may also use the satisfying assignment to look for improvements of
+the form $x = y$, where $x$ and $y$ are both variables.
+We look at the satisfying assignment, ignoring variables
+that were already improved to constants, and consider pairs of variables
+that happen to have the same value in this assignment. If $x$ and $y$
+are two such variables, then we try to prove that $x = y$ must hold
+under the current assumptions:
+\begin{Verbatim}
+mustEqual s x y =
+  solverProve s (eq (SMT.const x) (SMT.const y))
+\end{Verbatim}
+This process take $O(n^2)$ time, where $n$ is the number of variables,
+but we only need to call the solver for pairs of variables that have the same
+value---if they do not, then the current assignment provides a counter
+example indicating that the variables do not need to be equal in general.
+
+\paragraph{Improve Using a Linear Relation.}
+This improvement is a generalization of the previous example, that
+is specific to linear arithmetic.  The idea is to try to discover
+improving equations of the form $y = A * x + B$.  This type of
+improvement is a little different from the others, in that the right-hand
+side of the equation contains terms that are in the theory that we are solving.
+In general, such constraints are not very useful for the rest of GHC,
+as other parts of the constraint solver do not know about our theory
+(i.e., the functions \Verb"*" or \Verb"+") and, if we can prove it, then
+we must already know about this improvement!  This sort of improvement, is
+useful in one very common case, however:  if $y$ is a unification variable,
+then having this type of improving equation enables GHC to instantiate
+the variable, which leads to better type inference and simpler constraints,
+so we attempt this improvement only when the previous two have failed,
+and only for the unification variables in the constraints.
+
+We can compute the values for $A$ and $B$, if we have two examples for
+$x$ and two examples for $y$. Note that each call to the solver in
+the previous improvement steps gave rise to a potential additional set
+of examples, so we have plenty of points to work with: we take care to
+remember the values for the variables from two models.   It is also
+important that the $x$ values be different, but we can always find such
+examples: if we cannot find such an example, then $x$ would have only
+one possible value, and it would have been improved already.  It turns
+out that this sort of situations occur fairly frequently.  Consider,
+for example, the following program fragment:
+\begin{Verbatim}
+f :: Proxy (a + 1)
+f = Proxy
+
+g :: Proxy (b + 2)
+g = f
+\end{Verbatim}
+In this example, GHC needs to infer how to instantiate \Verb"f"'s type
+parameter, \Verb"a", and to do so, our plug-in needs to figure out that
+\Verb"a" may be instantiate to \Verb"b + 1".
+
+Once we compute candidate values for $A$ and $B$, we still need to
+invoke the solver to validate that the compute relation holds in all
+instances of the constraints, and not just the two that we happened to pick.
+
+Our current implementation only considers linear relations between pairs
+of variables as we were concerned about the performance penalty if
+we tried combinations of more variables: the math will work out, but
+the algorithm might become too slow to be practical, and alternative
+solutions based on rewriting might be more suitable.
+
+\paragraph{Custom Improving Rewrites.}
+All the improvements that we've defined so far work by {\em only} using
+the solver---note that we were just looking at satisfying assignments,
+and did not have any special rules about the shapes of the constraints.
+It also makes sense to extend the system with custom rewrite rules,
+which has the potential of speeding up performance, and also adding
+support for features that go beyond the solver's capabilities.
+We have experimented with various custom rules, but it is as yet
+unclear what constitutes a good set of rewrites.
+
+\subsection{Solving Constraints}
+
+The process of solving constraints is a straight-forward call the
+solver.  The only thing we need to do before solving constraints is
+to remove the assertions that were added during the consistency and
+improvement stages.  Recall that GHC solves implication constraints
+with two call to the constraint solver: the first one asserts
+the given constraints, while the second one actually solves goals.
+
+While we are asserting the assumption, we just check for consistency
+and improvement (to generate new given equalities), but we perform
+no solving, as there is nothing to solve.  During the solving stage,
+we mark the solver's state after we've asserted the givens,
+then we assert the new goals, check for consistency, and improvement,
+and then, before solving, revert back to the state where only
+the givens are asserted.
+\begin{Verbatim}
+solverSimplify s wanteds =
+  solverPrepare s wanteds $ \others our_wanteds ->
+  do res <- mapM tryToSolve our_wanteds
+     let (unsolved, solved) = partitionEithers res
+     return (solved, unsolved ++ others)
+  where
+  tryToSolve (ct,e) =
+    do proved <- solverProve s e
+       if proved
+         then return (Right ct)
+         else return (Left ct)
+\end{Verbatim}
+The call to \verb"solverPrepare" identifies the constraints that belong
+to our theory and translates them the SMTLIB language, where \Verb"others"
+are the constraints that are completely outside our theory, and
+\Verb"out_wanteds" are the constraints that we know about. Then,
+we invoke the solver for each goal, and see if we can prove it.
+
+\paragraph{Evidence.} When we solve a constraint, GHC expects some
+evidence to be produced, explaining why the constraint holds
+\cite{fc-evidence}.  The same is expected when generating new given
+constraints.  While this evidence is not used for code generation,
+it is a very useful for sanity checking while working with the constraint
+solver.  Unfortunately, at present, our plug-in does not produce any
+meaningful evidence, beyond indicating that the fact was produce
+by using the SMT solver. In principle, SMT solvers should be able to
+produce a proof, when they have concluded that a set of assertions
+is unsatisfiable.  Unfortunately, this is not a commonly used feature,
+so many provers do not support it out of the box, and the format of the
+proofs is not standardized.
 
@@
 
-\paragraph{Integers.}
-\paragraph{Boolean.}  SAT solver in the type system
-\paragraph{Bit-vectors.}  Aribtrary modulo arithmeitc; finite sets of effects.
-\paragraph{Sets.}
+In the previous Section, we described the core working of a procedure
+for integrating an SMT solver with GHC's constraint solver.  While
+we concentrated on the theory of natural numbers and linear arithmetic,
+little of the algorithm is specific to that theory.
+
+It would be fairly easy to add support for other theories, simply
+by declaring additional ``uninterpreted'' kinds and type-functions
+in Haskell.  In the rest of this Section we explore some of these
+possibilities.
+
+\subsection{A Theory for Booleans}
+We already made use of the lifted \Verb"Bool" type, but so far Haskell
+does not support any interesting operations, beyond the \Verb"<=?" relation
+that we described.  Programmers are already experimenting with defining
+type functions. For example, with the current version of GHC we may
+define a function for conjunction of type-level booleans:
+\begin{Verbatim}
+type family And a b where
+  And False x = False
+  And True x  = x
+\end{Verbatim}
+
+Definitions like this work for simple evaluation, but do not support
+more sophisticated reasoning.  For example, if GHC encounters
+the constraint \Verb"And x y ~ True", it will not be able to conclude
+that the only way to solve this is if \Verb"x ~ True" and \Verb"y ~ True"
+both hold.  Using the approach outlined in the previous Section, we
+gain the full power of a SAT solver in the type system---SMT solvers
+can reason about booleans, indeed a SAT solver is usually an important
+part of the implementation of an SMT solver.
+
+A good starting point for the theory of booleans, would be the following
+signature:
+\begin{Verbatim}
+type family And (a :: Bool) (b :: Bool) :: Bool
+type family Or  (a :: Bool) (b :: Bool) :: Bool
+type family Not (a :: Bool)             :: Bool
+\end{Verbatim}
+
+\subsection{A Theory for Integers}
+
+So far, our work uses natural numbers, which are quite useful for
+keeping track of the sizes of data-structures such as arrays, vectors,
+lists, bit-vectors, etc.  The standard theory supported by the SMT solvers
+is actually linear arithmetic over the integers---in our implementation
+we have to do some extra work to assert that we are only interested
+in non-negative solutions.
+
+The main challenge for supporting both integers and natural numbers
+at the type level is not the technology for solving the equations, but
+the notation!  One possible signature for the theory of type-level
+integers is:
+\begin{Verbatim}
+type family ToInt  (a :: Nat) :: Int
+type family Negate (a :: Int) :: Int
+type family Add    (a :: Int) (b :: Int) :: Int
+type family Sub    (a :: Int) (b :: Int) :: Int
+type family Mul    (a :: Int) (b :: Int) :: Int
+\end{Verbatim}
+While this notation is somewhat verbose, at least it has the benefit
+of exposing the functionality to programmers, who may use a more suitable
+notation depending on their needs.  For example, if a program works
+mostly with integers, one could add the following declaration, to get
+the more usual mathematical notation:
+\begin{Verbatim}
+type (+) a b = Add a b
+\end{Verbatim}
+
+\subsection{A Theory of Bit-Vectors}
+SMT solvers also have good support for reasoning about fixed-length bit-vectors.
+Unlike the theory of linear arithmetic, the theory of bit-vectors
+supports arbitrary multiplication and division, as well as various bit-wise
+operations, similar to the operations available in hardware.
+
+\paragraph{Tracking Sizes.}
+If we are interested in keeping track of the sizes of data-structures
+in memory, using 64-bit bit-vectors is a reasonable compromise
+which provides more expressive reasoning than linear arithmetic,
+and accommodates virtually all data-structures that fit in a computer's memory.
+
+\paragraph{Finite Sets.}
+Another potential application for the theory of bit-vectors is
+to implement finite sets of integers at the type level.
+As is usual at the value level, each bit is used to indicating the presence
+or absence of an element in the set, and the bitwise \Verb"and"
+and \Verb"or" operations correspond to intersection and union.
+Such functionality would be useful in the implementation of various effect
+systems.
 

### 2015-03-13 docs/paper.tex
More paper

@@
 \usepackage{url}
+\usepackage[utf8]{inputenc}
 
@@
 
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
+\conferenceinfo{Haskell Symposium 2015}{September, 2015, Vancouver, Canada} 
 \copyrightyear{2015} 
@@
 
-\title{Title Text}
+\title{Improving Haskell Types with SMT}
 
@@
 \begin{abstract}
-We present a technique for integrating GHC's type-checker with the
-an SMT solver.  The technique was developed to add support for reasoning
+We present a technique for integrating GHC's type-checker with an SMT solver.
+The technique was developed to add support for reasoning
 about type-level functions on natural numbers, and so our implementation
-uses the theory of linear arithmetic.  The technique, however, is more
-general and makes it possible to experiment with other external decision
-procedures, such as reasoning about type-level booleans, bit-vectors,
-or any other theory supported by SMT solvers.
+uses the theory of linear arithmetic.  However, the approach is not
+limited to this theory, and makes it possible to experiment with other
+external decision procedures, such as reasoning about type-level booleans,
+bit-vectors, or any other theory supported by SMT solvers.
 \end{abstract}
@@
 
-XXX: Structure of the paper
+The paper is organized as follows:  we start with a brief overview
+of SMT solvers form a user's perspective (Section~\ref{smt}); this
+section may be skipped by readers familiar with the tools.  Then,
+in Section~\ref{GHC}, we introduce the basic concepts of GHC's constraints
+solver, which is necessary for putting the rest of the paper in context.
+Section~\ref{algorithm} contains the details of the algorithm for
+integrating an SMT solver with GHC, and Section~\ref{other} explains
+other theories that could be added to GHC using the same technique.
+Finally, Section~\ref{modular-typechecker} discusses the possibility
+of using the ideas in this paper, and previous work on the
+implementation of SMT solver, to engineer a modular constraint solver.
 
@@
 \section{Integrating GHC with an SMT Solver}
+\label{algorithm}
 
@@
 
+\paragraph{Natural Numbers.}
+We are interested in working with natural number, however, the SMT
+decision procedure works with integers.  To work around this mismatch,
+we are careful that whenever we declare an SMT variable corresponding
+to a type of kind \Verb"Nat", we also add a formula asserting that
+the variable is not negative.
+
+\subsection{Communication with the Solver}
+
+Different SMT solvers support different mechanisms for interacting
+with other programs.  The most efficient way to interact with a solver
+is to link it with the program, and call into the various methods,
+using whatever API is exported by the solver.  Unfortunately, this
+approach requires commitment to a specific solver as the low-level
+API of the solvers differ.
+
+In our work, we wanted to have the flexibility to experiment with
+different solvers, so we opted for a slightly less optimal but
+considerably more portable approach.  The multitude of SMT languages
+has been recognized as a problem by the SMT community, and the
+SMT-LIB standard \cite{smtlib2} was developed to address the issue.
+A fair number of solvers have support for the standard, so
+to communicate with the solver we developed a Haskell library
+\cite{simple-smt}, which can interact with an SMT solver process
+by using the SMTLIB language.  The library manages the connection
+to the solver, and also provides combinators to create terms
+in the common SMT theories.  In the future, we may add support for
+direct integration with an SMT solver, and the only impact
+should be improved performance.
+
+The code samples in the rest of this Section use the following
+API for interaction with the solver:
+\begin{Verbatim}
+data Solver :: *
+data Result = Sat | Unsat
+
+solverAssert :: Solver -> Expr -> IO ()
+solverCheck  :: Solver -> IO Result
+solverPush   :: Solver -> IO ()
+solverPop    :: Solver -> IO ()
+\end{Verbatim}
+These correspond directly to the commands used in the examples in
+Section~\ref{smt}.
+
 \subsection{Checking for Consistency}
@@
 solver if the result is satisfiable.  If this is not the case, then
-we report and error, otherwise we proceed with the algorithm.  Note that
-since we are working with generalized constraints, if we find an
+we report and error, otherwise we proceed with the algorithm.
+\begin{Verbatim}
+checkConsistent s cs =
+  do mapM_ (solverAssert s) cs
+     res <- solverCheck s
+     return (res == Sat)
+\end{Verbatim}
+
+Note that since we are working with generalized constraints, if we find an
 inconsistency, then we are sure that we've detected a problem.  However,
@@
 \section{Other Theories}
+\label{other}
 
@@
 
-Relation between type-checker and
-the Nelson-Oppen method for combining
-decision procedures.
+In Section~\ref{algorithm} we outlined an algorithm for integrating for
+using an SMT solver to solver a certain class of constraints, and then
+in Section~\ref{other} we observed that it is possible---and potentially
+useful---to solve many other constraints.  This raises an interesting
+question: is it possible to structure the entire constraint solver for
+a complex programming language, such as Haskell or ML, using a modular
+approach, similar to the structure of an SMT solver?
 
-\section {Conclusions}
+We do not know that answer to this question, but the rest of this Section,
+we present a little bit about the structure of SMT solvers, and point out
+similarities with GHC's constraint solver, which suggests that this might
+be a fruitful area for future research.
+
+\subsection{The Nelson-Oppen Method}
+
+An SMT solver uses a collection of independent decision procedures,
+and orchestrates them to compute an answer to a given query.  A well-known
+technique for combing decision procedures is due to Nelson and Oppen
+\cite{NelsonOppen}.  Their algorithm starts with a collection of
+orthogonal theories (i.e., the symbols in the theories do not overlap),
+and shows how we may simplify terms in the combined language of the
+theories using {\em equality propagation}.  The equality propagation procedure
+has a lot of similarities to the algorithm presented in this paper,
+and this is entirely by design on our part.
+
+First, constraints that mention symbols from multiple theories are
+rewritten so that each constraint mentions only symbols in a single theory.
+This is done by naming ``foreign'' sub-terms, in the same way we described
+earlier in the paper.
+
+Each collection of constraints is sent to the appropriate decision
+procedure to be checked for consistency.  If one of the procedures
+detects an inconsistency, then we know that the whole collection of
+constraints is inconsistent and we may stop.
+
+If the constraints look consistent, then the decision procedures
+are given a chance to ``communicate'' with each other via equality
+constraints.  This is analogous to the improvement step in what we
+described.  A key observation of Nelson and Oppen is that the only
+constraints that need to be communicated are equalities between variables.
+The intuition behind this observation is that other equalities would
+feature symbols that are ``foreign'' to the other decision procedures,
+and they would not know what to do with them.
+
+So, if one decision procedure emits an equality constraint that was not
+known by some of the others, we propagate this information, and again
+check for consistency and further improvement.
+
+Finally, the Nelson-Oppen technique also allows for a more general
+form of communication between procedures---a decision procedure may
+indicate that it can solve a problem in multiple ways, by emitting
+a {\em disjunction} of equality constraints.  If this happens,
+then the coordinating logic has to explore all possibilities, the problem being
+satisfiable if any one of the options succeeds, and it is not satisfiable
+otherwise.  Searching the options introduced by disjunctions is typically
+implemented with the aid of a SAT solver, which is usually considerably
+more efficient than simply trying all possibilities, as SAT solvers
+propagate information to prune the search space.
+
+To implement the backtracking search efficiently, the decision procedures
+are typically incremental (i.e, they can revert back to a previous
+known state).  Some techniques can improve this search process further,
+by using decision procedures that also produce proofs, \cite{cvc4-lin-arith},
+as these proofs can be used to identify dependencies between variables.
+
+So, the overall setups is that we have a collection of independent
+decision procedures---each specializing in a single theory---and
+they communicate by disjunctions of equalities.  There are three cases,
+depending on the shape of the disjunction:
+\begin{itemize}
+\item an empty disjunction indicates a conflict,
+\item a singleton disjunction is a certain improvement, and we propagate
+it across theories,
+\item a proper disjunction results in search.
+\end{itemize}
+
+\subsection{Theories of Haskell}
+
+GHC's constraint solver is not structured exactly in the same way as outlined
+in the Nelson-Oppen simplification procedure, there are many similarities.
+In particular, equalities play a central role in disseminating information,
+but GHC currently does not do any back-tracking search (i.e., decision
+procedure may only report inconsistency, or certain knowledge).  However,
+there are situations where backtracking search is potentially useful.
+For example, when solving class constraints in the context of overlapping
+instances, or instance chains \cite{instance-chains}, it is useful
+to consider the context of instances, but back-track and try a different
+instance, if the context is found to be unsatisfiable.
+
+GHC's implementation of Haskell contains numerous extensions, and so
+there are many candidates for potential ``theories'', that is, constraints
+that are solved by a dedicated decision procedure.  Engineering a
+constraint solver in this style would be very beneficial, both
+as one could reason about the correctness of each decision separately,
+but also, because it would make it simpler to experiment with new
+extensions to the system.
+
+\section {Conclusion}
+
+The paper introduces the basics of GHC's constraint solver, and shows
+how to connect it with an external decision procedure that produces
+satisfying assignments.  The initial motiviation for this work was
+to improve the type-level support for natural numbers, but the solution
+turned out to be more general, and opens up the door for intersting
+new extentions to Haskell's type system.  More generally, we have
+identified similarities between the techniques used to implement SMT solvers,
+and the implementaion of GHC's constraint solver, which suggests
+the possibility of a modular constraint solver design.
 

### 2015-03-13 src/TypeNatSolver.hs
Filter out deriveds that also occur in givens.
Fixes #1.

@@ -145,7 +145,7 @@ solverEntry s givens derived wanteds =
 
        TcPluginOk [] new_cts ->
           do (solved,_) <- solverSimplify s wanteds
-             let known = map ctPred (derived ++ wanteds)
+             let known = map ctPred (givens ++ derived ++ wanteds)
              return (TcPluginOk solved (filter (ctNotMember known) new_cts))
 
        TcPluginOk _ _ -> panic "solveImprove returned Solved!"

### 2015-03-13 docs/paper.tex
Add some examples.

@@
 
-The paper is organized as follows:  we start with a brief overview
+\subsection{Examples}
+
+We illustrate the utility of the functionality provided by our
+algorithm with a few short examples.  A very common example in
+this area is to define a family of singleton types that links
+type-level natural numbers to run-time constants that represent
+them:
+\begin{Verbatim}
+data UNat :: Nat -> * where
+  Zero :: UNat 0
+  Succ :: UNat n -> UNat (n + 1)
+\end{Verbatim}
+Here we've used a unary representation of the natural numbers,
+and each member of the family, \Verb"UNat n", has exactly one
+inhabitant, namely the natural number \Verb"n" represented in
+unary form.  Because we are using a GADT, we can pattern match
+on the constructors of the type and gradually learn additional
+information about the value being examined.
+
+Next, we define a function to add two such numbers:
+\begin{Verbatim}
+uAdd :: UNat m -> UNat n -> UNat (m + n)
+uAdd Zero x      = x
+uAdd (Succ x) y  = Succ (uAdd x y)
+\end{Verbatim}
+While this is a simple definition, and we are unlikely
+to have gotten it wrong, it is nice to know that GHC
+is checking our work!  Had we made a mistake, for
+example, by mis-typing the recursive call as \Verb"uAdd x x",
+we would get a type error:
+\begin{Verbatim}
+  Could not deduce (((n1 + n1) + 1) ~ (m + n))
+    from the context (m ~ (n1 + 1))
+\end{Verbatim}
+It is a small detail, but it is worth pointing out that
+GHC would have been just as happy had we defined the type
+of \Verb"Succ" like this:
+\begin{Verbatim}
+  Succ :: UNat n -> UNat (1 + n)
+\end{Verbatim}
+The difference is in the return type, in the one case it is
+\Verb"n + 1", and in the other it is \Verb"1 + n".  While
+to a human this looks like an insignificant difference,
+in many systems it is significantly easier to work
+with the one definition, but not the other.
+
+The unary natural numbers are handy if we are planning to
+do iteration.  However, if we are working on some sort of
+divide-and-conquer algorithm, we often need to split the
+input in two.  In this situation, a different family of
+singletons is more useful:
+\begin{Verbatim}
+data BNat :: Nat -> * where
+  Empty :: BNat 0
+  Even  :: (1 <= n) => BNat n -> BNat (2 * n)
+  Odd   :: BNat n -> BNat (2 * n + 1)
+\end{Verbatim}
+This is another singleton family, where the type \Verb"BNat n"
+is inhabited by a single member, \Verb"n".  However, in this
+case we learn different information by pattern examining
+the values:  an \Verb"Empty" value may not be split,
+an \Verb"Even" value may be split into two equal non-empty
+parts, while an \Verb"Odd" value may be split into two parts,
+and there will be one element left over.
+
+So, how do we add such numbers?  There are more cases to
+consider:
+\begin{Verbatim}
+bAdd :: BNat m -> BNat n -> BNat (m + n)
+bAdd Empty x            = x
+bAdd x Empty            = x
+bAdd (Even x) (Even y)  = Even (bAdd x y)
+bAdd (Even x) (Odd y)   = Odd  (bAdd x y)
+bAdd (Odd x)  (Even y)  = Odd  (bAdd x y)
+bAdd (Odd x)  (Odd  y)  = Even (bSucc (bAdd x y))
+\end{Verbatim}
+The correctness of this definition is much less obvious,
+and we did make a couple of mistakes before getting it
+right!  In the last case, we use the auxiliary function
+\Verb"bSucc", which increments a binary natural number
+by one:
+\begin{Verbatim}
+bSucc :: BNat m -> BNat (m + 1)
+bSucc Empty     = Odd Empty
+bSucc (Even x)  = Odd x
+bSucc (Odd x)   = Even (bSucc x)
+\end{Verbatim}
+
+While these examples are somewhat simplistic, we hope
+that they demonstrate the utility of the technology,
+and show that using the extension feels quite natural
+to a Haskell programmer.
+
+\subsection{Structure of the Paper}
+
+The rest of the paper is organized as follows:
+we start with a brief overview
 of SMT solvers form a user's perspective (Section~\ref{smt}); this

### 2015-03-13 docs/Examples.hs

@@ -1,0 +1,39 @@
+{-# OPTIONS_GHC -fplugin=TypeNatSolver #-}
+{-# LANGUAGE DataKinds #-}
+{-# LANGUAGE TypeOperators #-}
+{-# LANGUAGE TypeFamilies #-}
+{-# LANGUAGE GADTs #-}
+{-# LANGUAGE KindSignatures #-}
+module Examples where
+
+import GHC.TypeLits
+import Data.Proxy
+
+data UNat :: Nat -> * where
+  Zero :: UNat 0
+  Succ :: UNat n -> UNat (n + 1)
+
+uAdd :: UNat m -> UNat n -> UNat (m + n)
+uAdd Zero x      = x
+uAdd (Succ x) y  = Succ (uAdd x y)
+
+
+data BNat :: Nat -> * where
+  Empty :: BNat 0
+  Even  :: (1 <= n) => BNat n -> BNat (2 * n)
+  Odd   :: BNat n -> BNat (2 * n + 1)
+
+bSucc :: BNat m -> BNat (m + 1)
+bSucc Empty     = Odd Empty
+bSucc (Even x)  = Odd x
+bSucc (Odd x)   = Even (bSucc x)
+
+bAdd :: BNat m -> BNat n -> BNat (m + n)
+bAdd Empty x            = x
+bAdd x Empty            = x
+bAdd (Even x) (Even y)  = Even (bAdd x y)
+bAdd (Even x) (Odd y)   = Odd  (bAdd x y)
+bAdd (Odd x)  (Even y)  = Odd  (bAdd x y)
+bAdd (Odd x)  (Odd  y)  = Even (bSucc (bAdd x y))
+
+

### 2015-03-14 docs/paper.tex
Typos and clarifications.

@@
 \cite{Eisenberg2012,Lindley2013,Eisenberg2014}.  The additional
-expressiveness enables Haskell programmers to express more
+expressiveness enables Haskell programmers to maintain more
 invariants at compile time, which makes it easier to develop
@@
 for embedding domain specific languages \cite{ivory-experience}.
-The Haskell compiler is not just a translator, but it also
-serves as a tool that analyzes the program, and helps
-finds common mistakes early in the development cycle.
+The Haskell compiler is not just a translator from source code
+to executable binaries, but it also serves as a tool that
+analyzes the program and helps find common mistakes early
+in the development cycle.
 
@@
 on the constructors of the type and gradually learn additional
-information about the value being examined.
+information about the value being examined.  The kind \Verb"Nat"
+is inhabited by types corresponding to the natural number
+(e.g., \Verb"0", \Verb"1", \dots), and \Verb"(+)" is a type-level
+function for adding natural numbers.
 
@@
 The rest of the paper is organized as follows:
-we start with a brief overview
-of SMT solvers form a user's perspective (Section~\ref{smt}); this
-section may be skipped by readers familiar with the tools.  Then,
-in Section~\ref{GHC}, we introduce the basic concepts of GHC's constraints
+we start with a brief overview of SMT solvers form a user's perspective
+(Section~\ref{smt}).  Then, in Section~\ref{GHC}, we introduce the
+basic concepts of GHC's constraints
 solver, which is necessary for putting the rest of the paper in context.
@@
 constraints arise from the use of \Verb"0" and \Verb"(==)" respectively,
-and the given constraint is obtain by pattern matching with \Verb"EInt".
+and the given constraint is obtained by pattern matching with \Verb"EInt"
+because we know that the type parameter of \Verb"E" must be \Verb"Int".
 
@@
 \subsection{Type-Checker Plugins}
-In this paper we describing a fairly general extension to the constraint
-solver, however one could always imagine others too. Indeed, there is
-existing work that extends GHC with support for units of measure
-\cite{units-of-measure}.  Instead of having many ad-hoc extensions directly
+In this paper we describe a fairly general extension to the constraint
+solver, but other researchers are interested in different extensions,
+for example, to add support for units of measure \cite{units-of-measure}.
+Instead of having many
+ad-hoc extensions directly
 in GHC's constraint solver, we collaborated to define and implement an
@@
 Instead, derived equalities help with type inference, by guiding the
-instantiation of {\em unification variables} with types. In general,
+instantiation of {\em unification variables}. In general,
 it is always sound to instantiate a unification variable with whatever
@@
 plug-in, and so the input to the algorithm is a collection of constraints
-that GHC has determine to be inert. This means that GHC simplified everything
-as much as it could, improved using equalities, and basically cannot see
+that GHC has determined to be inert. This means that GHC simplified everything
+as much as it could, improved using equalities, and cannot see
 anything else to do.  The inert constraints are presented to the plug-in
-in three collections, based on the constraints' provenance: one group contains
+in three collections based on the constraints' provenance: one group contains
 the given constraints (i.e., assumptions), one group contains derived
@@
 In the current implementation, we consider equalities and inequalities
-between a subset of the types of kind \Verb"Nat".   The kind \Verb"Nat" is
+between a subset of the type-expressions of kind \Verb"Nat".
+The kind \Verb"Nat" is
 inhabited by an infinite family of type-constants:
@@
 complex reasoning---involving variables---to the algorithm being presently
-described.
+described.  Technically, the clean-up is not necessary because the
+plug-in will perform as much evaluation as it needs to solve the constraints,
+but it is a lot more efficient to simply evaluate known constants without
+consulting the external solver.  Also, at present there is no standard
+way to declare that these type function are handled by a special solver---%
+we could use a closed type family with no equations to prohibit user
+defined instances, but this is a bit of an abuse of an unrelated feature
+of the type system.
 
@@
 
-The constraints of interest to the algorithm are summarized by the following
-grammar:
+The constraints supported by the external solver are summarized by the
+following grammar:
 \begin{align*}
@@
 \paragraph{Natural Numbers.}
-We are interested in working with natural number, however, the SMT
+We are interested in working with natural numbers, however, the SMT
 decision procedure works with integers.  To work around this mismatch,
@@
 data Solver :: *
+data Expr   :: *
 data Result = Sat | Unsat
@@
 \begin{Verbatim}
+checkConsistent :: Solver -> [Expr] -> IO ()
 checkConsistent s cs =
@@
 \begin{Verbatim}
+minimize :: Solver ->
+            [Expr] -> -- part of conflict
+            [Expr] -> -- search for conflict
+            IO ()
 minimize s yes [] = return yes
@@
 
+minimize :: Solver ->
+            [Expr] -> -- part of conflict
+            [Expr] -> -- currently asserted
+            [Expr] -> -- search for conflict
+            IO ()
 search s yes mb (c : cs) =
@@
 \begin{Verbatim}
+solverProve :: Solver -> Expr -> IO Bool
 solverProve s p =
@@
 
+mustBeK :: Solver -> Name -> Value -> IO Bool
 mustBeK s x v =
-  solverPorve s (eq (SMT.const x) (value v))
+  solverProve s (eq (SMT.const x) (value v))
 \end{Verbatim}
@@
 \begin{Verbatim}
+mustEqual :: Solver -> Name -> Name -> IO Bool
 mustEqual s x y =
@@
 
-The process of solving constraints is a straight-forward call the
+The process of solving constraints is a straight-forward call to the
 solver.  The only thing we need to do before solving constraints is
@@
 improvement stages.  Recall that GHC solves implication constraints
-with two call to the constraint solver: the first one asserts
+with two calls to the constraint solver: the first one asserts
 the given constraints, while the second one actually solves goals.
@@
 and improvement (to generate new given equalities), but we perform
-no solving, as there is nothing to solve.  During the solving stage,
+no solving as there is nothing to solve.  During the solving stage,
 we mark the solver's state after we've asserted the givens,
@@
 \begin{Verbatim}
+solverSimplify :: Solver ->
+                  [Expr] -> IO ([Expr],[Expr])
 solverSimplify s wanteds =
@@
 We already made use of the lifted \Verb"Bool" type, but so far Haskell
-does not support any interesting operations, beyond the \Verb"<=?" relation
+does not support any interesting operations beyond the \Verb"<=?" relation
 that we described.  Programmers are already experimenting with defining
@@
 
-GHC's constraint solver is not structured exactly in the same way as outlined
-in the Nelson-Oppen simplification procedure, there are many similarities.
-In particular, equalities play a central role in disseminating information,
-but GHC currently does not do any back-tracking search (i.e., decision
-procedure may only report inconsistency, or certain knowledge).  However,
-there are situations where backtracking search is potentially useful.
-For example, when solving class constraints in the context of overlapping
-instances, or instance chains \cite{instance-chains}, it is useful
-to consider the context of instances, but back-track and try a different
-instance, if the context is found to be unsatisfiable.
+While GHC's constraint solver currently is structured quite differently
+from an SMT solver, the similarities---in particular, the use of equality
+constraints to disseminating information---suggest that there may be
+a more modular way to structuring the solver.
+
+While GHC currently does not do any back-tracking search (i.e., decision
+procedure may only report inconsistency, or certain knowledge), this
+is also the case for Nelson-Open if all theories are convex.  In addition,
+there are interesting cases where backtracking might be potentially useful
+in Haskell too.  For example, when solving class constraints in the
+context of overlapping instances, or instance chains \cite{instance-chains},
+it is useful to consider the context of instances,
+but back-track and try a different instance, if the context is found
+to be unsatisfiable.
 
@@
 constraint solver in this style would be very beneficial, both
-as one could reason about the correctness of each decision separately,
-but also, because it would make it simpler to experiment with new
-extensions to the system.
+as one could reason about the correctness of each decision procedure
+separately, but also, because it would make it simpler to experiment
+with new extensions to the system.
 

### 2015-06-05 src/TypeNatSolver.hs
Fix-up to work with latest version of SimpleSMT

@@ -57,19 +57,15 @@ thePlugin opts = TcPlugin
   , tcPluginStop  = pluginStop
   }
 
-verbose :: Bool
-verbose = False
+quiet :: Int
+quiet = 1
 
 pluginInit :: [CommandLineOption] -> TcPluginM S
 pluginInit opts = tcPluginIO $
   do -- XXX: Use `opts`
      let exe  = "cvc4"
          opts = [ "--incremental", "--lang=smtlib2" ]
-     doLog <- if verbose
-                then SMT.newLogger
-                else return SMT.Logger { SMT.logTab = return ()
-                                       , SMT.logUntab = return ()
-                                       , SMT.logMessage = \_ -> return () }
+     doLog <- SMT.newLogger quiet
      proc  <- SMT.newSolver exe opts (Just doLog)
 
      SMT.setLogic proc "QF_LIA"
@@ -782,14 +778,13 @@ solverProve' s e =
 -- We prefer this to `trace` because trace is too verbose.
 -- This should not be used in "release" versions.
 solverDebugFun :: S -> String -> TcPluginM a -> TcPluginM a
-solverDebugFun s x m
-  | verbose = do tcPluginIO (do SMT.logMessage l x
-                                SMT.logTab l)
-                 a <- m
-                 tcPluginIO (do SMT.logUntab l
-                                SMT.logMessage l ("end of " ++ x))
-                 return a
-  | otherwise = m
+solverDebugFun s x m =
+  do tcPluginIO (do SMT.logMessage l x
+                    SMT.logTab l)
+     a <- m
+     tcPluginIO (do SMT.logUntab l
+                    SMT.logMessage l ("end of " ++ x))
+     return a
   where
   l = dbgLogger s
 

### 2015-06-05 src/TypeNatSolver.hs
Fix to avoid non-termination due to changes in plug-in interface.
This avoids non-termination in some common cases.  For details,
see the comment just above `isVarEq`.

In short: if a plug-in generates something which is equivalent
(but not syntactically the same) to something already in the inert set,
and GHC notices that fact, we go into a loop, because the plug-in just
keeps generating the new fact, and GHC keeps ignoring it.

The current fix is a work around, but this should really be fixed in
TcInteract in GHC---it can let us know if it disregarded a fact, and
nothing else changed, the wa

@@ -4,6 +4,7 @@ import Type      ( PredType, Type, Kind, TyVar, eqType
                  , getTyVar_maybe, isNumLitTy, splitTyConApp_maybe
                  , getEqPredTys, mkTyConApp, mkNumLitTy, mkEqPred
                  , typeKind, classifyPredType, PredTree(..), EqRel(..)
+                 , getTyVar_maybe, getEqPredTys_maybe
                  )
 import TyCon     ( TyCon )
 import TcEvidence ( EvTerm(..), mkTcAxiomRuleCo )
@@ -36,6 +37,8 @@ import Outputable
 
 import           Data.Map ( Map )
 import qualified Data.Map as Map
+import           Data.Set ( Set )
+import qualified Data.Set as Set
 import           Data.IORef ( IORef, newIORef, readIORef, writeIORef
                             , modifyIORef', atomicModifyIORef' )
 import           Data.List ( find, partition )
@@ -115,17 +118,21 @@ pluginSolve s gs ds0 ws0 =
   ppCts cs = vcat (map ppr cs)
 
 
-ctNotMember :: [PredType] -> Ct -> Bool
-ctNotMember tys ct = isNothing (find (eqType ty) tys)
-  where ty = ctPred ct
+ctNotMember :: [Ct] -> Ct -> Bool
+ctNotMember cts = \ct ->
+  case isVarEq ct of
+    Just (x,y) -> repFor x /= repFor y
+    _          -> isNothing (find (eqType (ctPred ct)) tys)
+  where
+  tys     = map ctPred cts
+  repFor  = makeEqVars cts
 
 solverEntry :: S -> [Ct] -> [Ct] -> [Ct] -> TcPluginM TcPluginResult
 solverEntry s givens _ [] =
   do res <- solverImprove s True givens
      case res of
        TcPluginOk [] new ->
-         do let known = map ctPred givens
-            return (TcPluginOk [] (filter (ctNotMember known) new))
+         return (TcPluginOk [] (filter (ctNotMember givens) new))
        x -> return x
 
 solverEntry s givens derived wanteds =
@@ -141,10 +148,84 @@ solverEntry s givens derived wanteds =
 
        TcPluginOk [] new_cts ->
           do (solved,_) <- solverSimplify s wanteds
-             let known = map ctPred (givens ++ derived ++ wanteds)
+             let known = givens ++ derived ++ wanteds
[…]

### 2015-06-05 docs/paper.tex
Fix typos

@@
 uAdd :: UNat m -> UNat n -> UNat (m + n)
-uAdd Zero x      = x
+uAdd Zero     y  = y
 uAdd (Succ x) y  = Succ (uAdd x y)
@@
 The rest of the paper is organized as follows:
-we start with a brief overview of SMT solvers form a user's perspective
+we start with a brief overview of SMT solvers from a user's perspective
 (Section~\ref{smt}).  Then, in Section~\ref{GHC}, we introduce the
@@
 
-This Section contains an introduction to the core functionality
+This section contains an introduction to the core functionality
 of a typical SMT solver, and may be skipped by readers who are already
@@
 a universally quantified formula. The idea is that we use the SMT solver
-to look for a {\em counter-example} to the formula, and if on such
+to look for a {\em counter-example} to the formula, and if no such
 example exists, then we can conclude the formula is valid.
@@
 
-In this Section, we present relevant aspects of GHC's constraint solver.
+In this section, we present relevant aspects of GHC's constraint solver.
 The full details of the algorithm \cite{outsidein} are beyond the scope
@@
 and {\em interacted} with the solver's state.  In the process of interaction
-we may solve constraint, generate new work, or report impossible constraints.
+we may solve constraints, generate new work, or report impossible constraints.
 If nothing interesting happens, then we relocate the constraint to the
@@
 over the natural numbers, as having a concrete theory makes it easier to explain
-the process and illustrate it with examples.  In the next Section,
+the process and illustrate it with examples.  In the next section,
 we discuss how, and why, we might want to consider other theories also.
@@
 
-Recall from Section~\ref{smt}, than to prove something, we make sure
+Recall from Section~\ref{smt}, that to prove something, we make sure
 that there are no counter examples.  The Corollary states that if there
-are no counter examples to the abstracted constraint, than there will
+are no counter examples to the abstracted constraint, then there will
 be no counter-examples to the original constraint as well, and so our
@@
 
-The code samples in the rest of this Section use the following
+The code samples in the rest of this section use the following
 API for interaction with the solver:
@@
 solver if the result is satisfiable.  If this is not the case, then
-we report and error, otherwise we proceed with the algorithm.
+we report an error, otherwise we proceed with the algorithm.
 \begin{Verbatim}
@@
 solver.  Unfortunately, at present, our plug-in does not produce any
-meaningful evidence, beyond indicating that the fact was produce
+meaningful evidence, beyond indicating that the fact was produced
 by using the SMT solver. In principle, SMT solvers should be able to
@@
 
-In the previous Section, we described the core working of a procedure
+In the previous section, we described the core working of a procedure
 for integrating an SMT solver with GHC's constraint solver.  While
@@
 by declaring additional ``uninterpreted'' kinds and type-functions
-in Haskell.  In the rest of this Section we explore some of these
+in Haskell.  In the rest of this section we explore some of these
 possibilities.
@@
 that the only way to solve this is if \Verb"x ~ True" and \Verb"y ~ True"
-both hold.  Using the approach outlined in the previous Section, we
+both hold.  Using the approach outlined in the previous section, we
 gain the full power of a SAT solver in the type system---SMT solvers
@@
 
-In Section~\ref{algorithm} we outlined an algorithm for integrating for
-using an SMT solver to solver a certain class of constraints, and then
+In Section~\ref{algorithm} we outlined an algorithm for using an SMT solver
+to solver a certain class of constraints, and then
 in Section~\ref{other} we observed that it is possible---and potentially
@@
 
-We do not know that answer to this question, but the rest of this Section,
+We do not know that answer to this question, but the rest of this section,
 we present a little bit about the structure of SMT solvers, and point out
@@
 and orchestrates them to compute an answer to a given query.  A well-known
-technique for combing decision procedures is due to Nelson and Oppen
+technique for combining decision procedures is due to Nelson and Oppen
 \cite{NelsonOppen}.  Their algorithm starts with a collection of
@@
 from an SMT solver, the similarities---in particular, the use of equality
-constraints to disseminating information---suggest that there may be
+constraints to disseminate information---suggest that there may be
 a more modular way to structuring the solver.
@@
 how to connect it with an external decision procedure that produces
-satisfying assignments.  The initial motiviation for this work was
+satisfying assignments.  The initial motivation for this work was
 to improve the type-level support for natural numbers, but the solution
-turned out to be more general, and opens up the door for intersting
-new extentions to Haskell's type system.  More generally, we have
+turned out to be more general, and opens up the door for interesting
+new extensions to Haskell's type system.  More generally, we have
 identified similarities between the techniques used to implement SMT solvers,
-and the implementaion of GHC's constraint solver, which suggests
+and the implementation of GHC's constraint solver, which suggests
 the possibility of a modular constraint solver design.

### 2015-06-05 examples/A.hs
Add additional examples.

@@ -19,6 +19,12 @@ data Vec :: Nat -> * -> * where
   Nil :: Vec 0 a
   Cons :: a -> Vec n a -> Vec (n + 1) a
 
+vecHead :: Vec (n + 1) a -> a
+vecHead (Cons x _) = x
+
+vecZipWith :: (a -> b -> c) -> Vec n a -> Vec n b -> Vec n c
+vecZipWith f Nil         Nil         = Nil
+vecZipWith f (Cons x xs) (Cons y ys) = Cons (f x y) (vecZipWith f xs ys)
 
 append :: Vec m a -> Vec n a -> Vec (n + m) a
 append Nil ys = ys
@@ -36,6 +42,5 @@ f1 = g1
 
 g1 :: Proxy (1 + a) -> ()
 g1 _ = ()
-
 
 

### 2015-06-05 docs/paper.tex
Correct type signatures.

@@
 \begin{Verbatim}
-checkConsistent :: Solver -> [Expr] -> IO ()
+checkConsistent :: Solver -> [Expr] -> IO Bool
 checkConsistent s cs =
@@
             [Expr] -> -- search for conflict
-            IO ()
+            IO [Expr]
 minimize s yes [] = return yes
@@
 
-minimize :: Solver ->
+search :: Solver ->
           [Expr] -> -- part of conflict
@@
           [Expr] -> -- search for conflict
-            IO ()
+          IO [Expr]
+
 search s yes mb (c : cs) =
@@
        _ -> search s yes (c : mb) cs
+
+search s yes mb [] = error "Impossible!"
 \end{Verbatim}

### 2015-06-06 docs/paper.tex
Make changes suggested by reviewers.

@@
 
+The approach presented in this paper uses similar tools---namely
+SMT solvers---to Liquid Haskell \cite{liquid-haskell}. However,
+the design and overall goals of the two techniques are somewhat
+different and, indeed, it makes perfect sense to use both systems
+in a single project.  Liquid Haskell enables Haskell programmers
+to augment Haskell definitions with refinements of the types,
+which may be used to verify various properties of the program.
+The refinment language is separate from the language of Haskell
+types---it is quite expressive, and is really aimed at formal
+verification.  As such, a function will have its ordinary
+type and, in addition, a refinment type, which serves to specify
+some aspect of the behavior of the function.  The checks performed
+by Liquid Haskell are entirely separate from the ones performed
+by GHC's type checker.  This is quite different from the
+technuqe described in this paper, which shows how to integrate
+an SMT solver directly into GHC's type chcker.  The benefit of
+of extending GHC's type chcker is that we don't need to
+introduce a separate language for specifications, instead
+we may reuse types directly.  Ultimately, however, the two
+approaches are not at all mutually exclusive.  The next
+section gives a brief flavor of what can be achieved using
+the technieques from this paper.
+
 \subsection{Examples}
@@
 question: are there concrete values for the uninterpreted constants
-that make all asserted formulas true.  Here is an example, using
-the notation of the SMTLIB standard \cite{smtlib2}
+that make all asserted formulas true. The solver may respond in
+one of three ways:
+\begin{itemize}
+\item \Verb"sat", which means that a satisfying assignment exists,
+\item \Verb"unsat", which means that a satisfying assignment does not exist,
+\item \Verb"unknown", which means that the solver did not find an answer,
+      often this is due to reaching some sort of limit (e.g, the solver
+      used up too much time).
+\end{itemize}
+
+This example uses the notation of the SMTLIB standard \cite{smtlib2}:
 
@@
 for example, to add support for units of measure \cite{units-of-measure}.
-Instead of having many
-ad-hoc extensions directly
-in GHC's constraint solver, we collaborated to define and implement an
-API for extending GHC's functionality via {\em type-checker plug-ins}.
-At present, we are aware of two main
-users of this API---our work, and the work on units of measure. We
-hope, however, that this mechanism would be useful to other researchers too,
+
+Instead of having many ad-hoc extensions directly in GHC's constraint solver,
+we collaborated to define and implement an API for extending GHC's
+functionality via {\em type-checker plug-ins}.  At present, we are aware
+of three users of this API---our work, a related plug-in that can
+solve numerical constraints using custom rewrite rules
+\cite{typelits-normalise},
+and the work on units of measure \cite{units-of-measure}.
+
+While this infrastructure is still brand new, and not yet stable,
+we hope that this mechanism may be useful to other researchers too,
 as it makes it fairly easy to experiment with various extensions to
-GHC's constraint solver.
+GHC's constraint solver.  Since the framework is aimed at GHC developers,
+and is not intended for use in everyday Haskell programming, the plug-ins
+have a lot of power.  Indeed, it is not difficult to write a plug-in
+that would violate type-safety, in much the same way one could do this
+by modifying GHC's constraint solver directly.  The full details of
+the type-checker plug-ins API may be found in the GHC User's Guide
+\cite{ghc-manual}. Here we discuss only some
+of the design choices that we considered, as they have a direct
+bearing on the workings of our plug-in.
 
@@
 Finally, derived equalities are implied by the given and wanted constraints
-jointly, so they may not be used to rewrite given or wanted constraints
-directly, as doing so may lead to circular reasoning.
-
-Instead, derived equalities help with type inference, by guiding the
+jointly, so they may not be used in proofs directly,
+as doing so may lead to circular reasoning.
+
+Instead, derived equalities help with type inference, by {\em guiding} the
 instantiation of {\em unification variables}. In general,
@@
 rewrite \Verb"Eq x" to \Verb"Eq Int", which we would proceed to solve
-as usual.
+as usual.  This is bit subtle:  a derived equality is never used directly
+in a proof---instead it allows us to {\em change what we are trying to prove}
+by instantiating unification variables, which may help discharge all
+constraints that need solving.
 
@@
 which, in turn, can be discharged by the custom solver for the \Verb"KnownNat"
-class.  Such collaboration between different solvers is quite common.
+class\footnote{The class \Verb"KnownNat" is defined in module
+\Verb"GHC.TypeLits".  It is used to implement singleton types linking
+type-level literals to their run-time representations as \Verb"Integer".}.
+Such collaboration between different solvers is quite common.
 Interestingly, this looks a lot like the collaboration between decision
@@
 The step of naming foreign types is similar to the flattening step performed
-by GHC's constraint solver, the main difference being that GHC names every
+by GHC's constraint solver\footnote{GHC's ``flattening'' pass names the
+results of all type functions.  This is done before passing constraints
+to GHC's internal solver, but the process is reverted before passing the
+constraints to the external plug-ins. }.
+The main difference being that GHC names every
 call to a type-function, while we name only terms outside of the theory.
@@
 does not have this ability.  To work around this, we implemented
-a custom minimization algorithm which work as follows:
-\begin{Verbatim}
-minimize :: Solver ->
+a custom reduction algorithm which work as follows:
+\begin{Verbatim}
+reduce :: Solver ->
             [Expr] -> -- part of conflict
@@
             IO [Expr]
-minimize s yes [] = return yes
-minimize s yes mb =
+reduce s yes [] = return yes
+reduce s yes mb =
   do solverPush s
@@
             solverAssert s c
-            minimize s (c : yes) mb
+            reduce s (c : yes) mb
 
@@
 
-The minimization process performs in $O(n^2)$ time, where $n$ is the
+The reduction process performs in $O(n^2)$ time, where $n$ is the
 number of constraints.  This has not been a big problem,
-as this minimization happens only when we report errors and, typically,
+as this reduction happens only when we report errors and, typically,
 the number of constraints is fairly low.
+
+Also note that the reduction algorithm may not produce the {\em smallest}
+collection of constraints to cause a contradiction.  For example, consider
+the constraints $A,B,C$, and suppose that $A,B$ lead to a contradiction
+together, and also $C$ causes a contradiction on its own.  Now, depending
+on the order in which we process the constraints, we may either compute
+$C$ as the result of the reduction, or $A,B$.  This is OK---in this
+case the context happens to contain multiple errors, so the reduction
+algorithm will simply pick one of them to report.  Of course, a program
+would have to eventually fix both errors before the program is accepted.
+The important property of the algorithm is that the reduced set does
+not contain redundant constraints, in the sense that if we removed
+some of the constraints the resulting constraint set would be satisfiable.
 
@@
 We can compute the values for $A$ and $B$, if we have two examples for
-$x$ and two examples for $y$. Note that each call to the solver in
-the previous improvement steps gave rise to a potential additional set
-of examples, so we have plenty of points to work with: we take care to
-remember the values for the variables from two models.   It is also
+$x$ and two examples for $y$.  It is quite standard to ask the SMT
+solver for additional examples: this is done by adding additional constraints
+that force the models to differ.  In our setting we know that we can
+always get another such examples---if this was not the case, a variable
+would have exactly one possible values, and we would have improved
+it to a constant, as describe previously.  Furthermore, since we do the linear
+relation improvement last, we usually can avoid extra calls to the solver
+because by this point we typically already have two models where
+the variables differ.  It is also
 important that the $x$ values be different, but we can always find such
 examples: if we cannot find such an example, then $x$ would have only
-one possible value, and it would have been improved already.  It turns
-out that this sort of situations occur fairly frequently.  Consider,
-for example, the following program fragment:
+one possible value, and it would have been improved already.
+
+We admit that this improvement is certainly somewhat bit ad-hoc: after all,
+why look for relations between two variables and more?  We implemented
+this improvement, because while experimenting with the implementation,
+we noticed that this sort of improvement is needed quite often,
+at least for a certain class of problems.  So, as far as we know, this
+simply a useful heuristic that is worth considering.  It is possible
+that alternative techniques (e.g., custom rewrite rules) could achieve
+the same effect more directly, and we are certainly considering exploring
+such approaches in future work.
+
+To see when this improvement is needed, consider the following program
+fragment:
 \begin{Verbatim}
@@
 
+\subsection{Our Implementation}
+\label{implementation}
+
+To demonstrate the feasibility of the ideas described in this section,
+we implemented a GHC plug-in, available from GitHub \cite{type-nats-solver}.
+The implementation is intended as a proof-of-concept, and is not
+yet ``industrial strength''---more engineering work would be required
+to make it more robust and easier to install and configure.
+
+The majority of the effort so far has been focused on understanding how
+to integrate an SMT solver with the type-checker, and we have not
+conducted extensive testing and measurement of the performance of the plug-in.
+It is clear that calling out to an external tool has a cost associated
+with it.  However, the preliminary results are encouraging---the plug-in
+functions quite well on small examples, but certainly more testing and
+experience is needed to find out if the approach will scale to larger
+projects.
+
+If despite all the disclaimers, the reader is still interested in
+experimenting with the plug-in, we provide the list of GHC options and
+extensions that we used in the examples from this paper:
+
+\begin{Verbatim}
+{-# OPTIONS_GHC -fplugin=TypeNatSolver #-}
+{-# LANGUAGE DataKinds #-}
+{-# LANGUAGE TypeOperators #-}
+{-# LANGUAGE TypeFamilies #-}
+{-# LANGUAGE KindSignatures #-}
+{-# LANGUAGE GADTs #-}
+\end{Verbatim}
+
+The option \Verb"fplugin=TypeNatSolver" instructs GHC to load the SMT solver
+plug-in.  The rest are Haskell extensions that are commonly used in this
+setting:
+\begin{itemize}
+\item \Verb"DataKinds" enables natural number literals at the type
+level.
+\item \Verb"TypeOperators" and \Verb"TypeFamilies" are need to work
+with type-level functions such as \Verb"+".
+\item \Verb"KindSignatures" is used to specify the kinds of type variables,
+for example, to state that some type-variable is of kind \Verb"Nat".
+\item \Verb"GADTs" enables Generalized Algebraic Data-types;  strictly
+speaking, these are not required to experiment with the plug-in, however,
+many of the more interesting uses of the plug-in end up using GADTs.
+\end{itemize}
+
+For the experiments in this paper, we used CVC4, version 1.4.  There was
+no deep reason for using this particular solver, other that when we started
+working on the project this was the solver with the least restrictive license.
+To communicate with the solver, plug-in uses the SMTLIB format (version 2).
+Many other SMT solvers support this format (e.g., Z3, Yices), so they
+should work just as well.
+
 \section{Other Theories}
@@
 
-It would be fairly easy to add support for other theories, simply
-by declaring additional ``uninterpreted'' kinds and type-functions
-in Haskell.  In the rest of this section we explore some of these
-possibilities.
+In principle, it should be fairly easy to add support for other theories,
+simply by declaring additional ``uninterpreted'' kinds and type-functions
+in Haskell.  Of course---as is often the case---doing so is not completely
+trivial, which is why we have not done so yet.  For example, while we
+have presented a mechanism for working with the symbols of a new theory
+in the type-checker, one still needs to make decisions about which symbols
+to import, or what notation to use.  In the rest of this section we
+explore some of these possibilities.
 
@@
 procedure may only report inconsistency, or certain knowledge), this
-is also the case for Nelson-Open if all theories are convex.  In addition,
+is also the case for Nelson-Open if all theories are convex \cite{NelsonOppen}.
+In addition,
 there are interesting cases where backtracking might be potentially useful

### 2015-07-18 docs/paper.tex
Fix typos

@@
 which may be used to verify various properties of the program.
-The refinment language is separate from the language of Haskell
+The refinement language is separate from the language of Haskell
 types---it is quite expressive, and is really aimed at formal
 verification.  As such, a function will have its ordinary
-type and, in addition, a refinment type, which serves to specify
+type and, in addition, a refinement type, which serves to specify
 some aspect of the behavior of the function.  The checks performed
@@
 by GHC's type checker.  This is quite different from the
-technuqe described in this paper, which shows how to integrate
-an SMT solver directly into GHC's type chcker.  The benefit of
-of extending GHC's type chcker is that we don't need to
+technique described in this paper, which shows how to integrate
+an SMT solver directly into GHC's type checker.  The benefit of
+of extending GHC's type checker is that we don't need to
 introduce a separate language for specifications, instead
@@
 section gives a brief flavor of what can be achieved using
-the technieques from this paper.
+the techniques from this paper.
 
@@
 Later on, if we need to return results to the core GHC solver that mention
-\Verb"x", we replace it by the orignal expression, \Verb"F 3".
+\Verb"x", we replace it by the original expression, \Verb"F 3".
 

### 2015-07-18 examples/A.hs
Add more examples, and a tricky bit that does not work.
We end up in this situation:

(2 <= a) => (a ~ ?n + 1)

This is interesting because while we know that ?n will always exist,
we don't know *what* it will be.  If we used subtraction, we could
define `?n ~ (a - 1)`, which in this case is safe because we know that
a is at least two.

We can start doing that, once the story with partial functions, such
as subtraction, clears up a bit more.

@@ -44,3 +44,20 @@ g1 :: Proxy (1 + a) -> ()
 g1 _ = ()
 
 
+
+vecMaximum :: Vec (n + 1) Int -> Int
+vecMaximum (Cons x xs) = go x xs
+  where
+  go :: Int -> Vec a Int -> Int
+  go m Nil         = m
+  go m (Cons n xs) = go (max m n) xs
+
+testLin :: Vec (2 + m) a -> a
+testLin xs = vecHead xs
+
+
+{-
+-- XXX: Tricky, requires us to "invent" a new variable.
+testLin :: (2 <= a) => Vec a Int -> Int
+testLin xs = vecMaximum xs
+-}

### 2015-07-18 docs/paper.tex
Improvements based on reviewer feedback (and fix page breaks)

@@
 \end{Verbatim}
-It is a small detail, but it is worth pointing out that
-GHC would have been just as happy had we defined the type
+It is a small detail, but it is worth pointing out that our
+GHC extension would have been just as happy had we defined the type
 of \Verb"Succ" like this:
@@
 
-While these examples are somewhat simplistic, we hope
-that they demonstrate the utility of the technology,
-and show that using the extension feels quite natural
-to a Haskell programmer.
+Note that while these examples are somewhat simplistic,
+they demonstrate the utility of the technology, and show that using
+the extension feels quite natural to a Haskell programmer.
+
+A more conventional example is the type of fixed-length vectors:
+\begin{Verbatim}
+data Vec :: Nat -> * -> * where
+  Nil  :: Vec 0 a
+  Cons :: a -> Vec n a -> Vec (n + 1) a
+\end{Verbatim}
+Note that this is very similar to \Verb"UNat", except that it also
+stores a value in each \Verb"Cons" node.  The operation of appending
+two vectors together follows exactly the same pattern as \Verb"uAdd".
+An example of a function that is meaningful for vectors, bit no
+so much for the unary natural numbers is reversing a vector, which
+may be defined as follows using our extension:
+\begin{Verbatim}
+vecReverse = go Nil
+  where
+  go :: Vec m a -> Vec n a -> Vec (m + n) a
+  go xs Nil = xs
+  go xs (Cons y ys) = go (Cons y xs) ys
+\end{Verbatim}
+This example also illustrates that our extension does not change any type
+signature requirements beyond what is already in GHC: the helper
+function \Verb"go" needs a type signature, because it pattern-matches
+using GADT constructors, however the type of \Verb"reverse" can be
+inferred as usual:
+\begin{Verbatim}
+*Example> :type vecReverse
+vecReverse :: Vec n a -> Vec n a
+\end{Verbatim}
 
@@
 
+\break
+
 \begin{Verbatim}
@@
 are:
+\break
 \begin{Verbatim}
@@
 
-We admit that this improvement is certainly somewhat bit ad-hoc: after all,
-why look for relations between two variables and more?  We implemented
-this improvement, because while experimenting with the implementation,
-we noticed that this sort of improvement is needed quite often,
-at least for a certain class of problems.  So, as far as we know, this
-simply a useful heuristic that is worth considering.  It is possible
-that alternative techniques (e.g., custom rewrite rules) could achieve
-the same effect more directly, and we are certainly considering exploring
-such approaches in future work.
-
+Once we compute candidate values for $A$ and $B$, we still need to
+invoke the solver to validate that the computed relation holds in all
+instances of the constraints, and not just the two that we happened to pick.
+
+\break
 To see when this improvement is needed, consider the following program
@@
 \end{Verbatim}
-In this example, GHC needs to infer how to instantiate \Verb"f"'s type
-parameter, \Verb"a", and to do so, our plug-in needs to figure out that
-\Verb"a" may be instantiate to \Verb"b + 1".
-
-Once we compute candidate values for $A$ and $B$, we still need to
-invoke the solver to validate that the compute relation holds in all
-instances of the constraints, and not just the two that we happened to pick.
+To check this program, GHC needs to solve the constraint \\
+\Verb"(?a + 1) ~ (b + 2)". The question mark in \Verb"?a" emphasizes that
+this is a unification variable, standing-in for a type that needs to be
+inferred. We cannot discharge this constraint immediately, because it
+does not hold {\em for all} possible values of \Verb"?a" and \Verb"b".
+However, since \Verb"?a" is a unification variable, we have the option
+of instantiating it to change the constraint.  Of course, we have to
+be careful that in doing so we do not loose any generality and, in this
+case, the safe instantiation is \Verb"?a := b + 1".  The purpose of the
+linear-relation improvement is to discover these kinds of
+instantiations.
+
+At first sight, this example may appear artificial, however it is an
+abstraction of a very common pattern: namely when one function calls
+another, and the type of the provided argument does not match
+syntactically the type of the expected argument.  As another example,
+consider the function \Verb"vecHead", which requires that its parameter
+has at least one element:
+\begin{Verbatim}
+vecHead :: Vec (n + 1) a -> a
+vecHead (Cons x _) = x
+\end{Verbatim}
+Now, if we tried to call \Verb"vecHead" with an expression of type
+\Verb"Vec (b + 2) a", we'd end up in exactly the same situation as the
+previous example, where we'd have to infer that \Verb"?n = b + 1".
 
@@
 solutions based on rewriting might be more suitable.
+
+We admit that this improvement is somewhat ad-hoc, but as the previous
+examples illustrate, it helps in situations that are quite common in
+practice.  It was also fairly simple to implement using the solver.
 
@@
 We have experimented with various custom rules, but it is as yet
-unclear what constitutes a good set of rewrites.
+unclear what constitutes a good set of rewrites.  Concrete custom
+rules that would help are various normalization rules, for example
+applying distributivity repeatedly, and also rules that rewrite
+equations so that there is just a unification variable on the one
+side of the equation, which would help it get instantiated.
 
@@
 setting:
+
+\break
 \begin{itemize}

### 2015-07-18 docs/paper.tex
Embed the bibliography

@@
 
-\bibliographystyle{abbrvnat}
-\bibliography{refs}
-
-% 
-% % The bibliography should be embedded for final submission.
-% 
-% \begin{thebibliography}{}
-% \softraggedright
-% 
-% \bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-% P. Q. Smith, and X. Y. Jones. ...reference text...
-% 
-% \end{thebibliography}
+% \bibliographystyle{abbrvnat}
+% \bibliography{refs}
+
+% The bibliography should be embedded for final submission.
+
+\begin{thebibliography}{}
+\softraggedright
+
+\providecommand{\natexlab}[1]{#1}
+\providecommand{\url}[1]{\texttt{#1}}
+\expandafter\ifx\csname urlstyle\endcsname\relax
+  \providecommand{\doi}[1]{doi: #1}\else
+  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi
+
+\bibitem[yic(2015)]{yices}
+\emph{Yices Manual}, 2015.
+\newblock URL \url{http://yices.csl.sri.com/papers/manual.pdf}.
+
+\bibitem[{Adam Gundry}(2015)]{units-of-measure}
+{Adam Gundry}.
+\newblock A typechecker plugin for units of measure.
+\newblock Submitted for publication., 2015.
+
+\bibitem[{Christiaan Baaij}(2015)]{typelits-normalise}
+{Christiaan Baaij}.
+\newblock A plug-in for normalising type literals in ghc.
+\newblock 2015.
+\newblock URL \url{https://github.com/clash-lang/ghc-typelits-natnormalise}.
+
+\bibitem[{Clark Barrett} et~al.(2010){Clark Barrett}, {Aaron Stump}, and
+  {Cesare Tinelli}]{smtlib2}
+{Clark Barrett}, {Aaron Stump}, and {Cesare Tinelli}.
+\newblock The {SMT-LIB} standard: Version 2.0.
+\newblock Technical report, Department of Computer Science, The University of
+  Iowa, 2010.
+\newblock URL
+  \url{http://smtlib.cs.uiowa.edu/papers/smt-lib-reference-v2.0-r12.09.09.pdf}.
+
+\bibitem[{Clark Barrett} et~al.(2011){Clark Barrett}, {Christopher L. Conway},
+  {Morgan Deters}, {Liana Hadarean}, {Dejan Jovanovi\'{c}}, {Tim King}, {Andrew
+  Reynolds}, and {Cesare Tinelli}]{cvc4}
+{Clark Barrett}, {Christopher L. Conway}, {Morgan Deters}, {Liana Hadarean},
+  {Dejan Jovanovi\'{c}}, {Tim King}, {Andrew Reynolds}, and {Cesare Tinelli}.
+\newblock {CVC4}.
+\newblock In \emph{Proceedings of the 23rd International Conference on Computer
+  Aided Verification}, Snowbird, Utah, USA, 2011.
+\newblock URL \url{http://dl.acm.org/citation.cfm?id=2032305.2032319}.
+
+\bibitem[{Dimitrios Vytiniotis} and {Simon Peyton Jones}(2013)]{fc-evidence}
+{Dimitrios Vytiniotis} and {Simon Peyton Jones}.
+\newblock Evidence normalization in System FC.
+\newblock In \emph{24th International Conference on Rewriting Techniques and
+  Applications}, Eindhoven, Netherlands, 2013.
+\newblock URL
+  \url{http://research.microsoft.com/en-us/um/people/simonpj/papers/ext-f/fc-new-tyco.pdf}.
+
+\bibitem[{Dimitrios Vytiniotis} et~al.(2011){Dimitrios Vytiniotis}, {Simon
+  Peyton Jones}, {Tom Schrijvers}, and {Martin Sulzmann}]{outsidein}
+{Dimitrios Vytiniotis}, {Simon Peyton Jones}, {Tom Schrijvers}, and {Martin
+  Sulzmann}.
+\newblock {OutsideIn(X)}: Modular type inference with local assumptions.
+\newblock \emph{Journal of Functional Programming}, September 2011.
+\newblock URL
+  \url{http://research.microsoft.com/apps/pubs/default.aspx?id=162516}.
+
+\bibitem[{Greg Nelson} and {Derek C. Oppen}(1979)]{NelsonOppen}
+{Greg Nelson} and {Derek C. Oppen}.
+\newblock Simplification by cooperating decision procedures.
+\newblock \emph{ACM Transactions on Programming Languages and Systems},
+  1\penalty0 (2), October 1979.
+\newblock URL \url{http://doi.acm.org/10.1145/357073.357079}.
+
+\bibitem[{Iavor S. Diatchki}(2015{\natexlab{a}})]{simple-smt}
+{Iavor S. Diatchki}.
+\newblock \emph{simple-smt: a {Haskell} library for working with SMT solvers},
+  2015{\natexlab{a}}.
+\newblock URL \url{http://hackage.haskell.org/package/simple-smt}.
+
+\bibitem[{Iavor S. Diatchki}(2015{\natexlab{b}})]{type-nats-solver}
+{Iavor S. Diatchki}.
+\newblock A plug-in for solving linear constraints using an smt solver.
+\newblock 2015{\natexlab{b}}.
+\newblock URL \url{https://github.com/yav/type-nat-solver}.
+
+\bibitem[{J. Garrett Morris} and {Mark P. Jones}(2010)]{instance-chains}
+{J. Garrett Morris} and {Mark P. Jones}.
+\newblock Instance chains: Type class programming without overlapping
+  instances.
+\newblock In \emph{Proceedings of the 15th ACM SIGPLAN International Conference
+  on Functional Programming (ICFP '10)}, Baltimore, Maryland, USA, 2010.
+\newblock URL \url{http://web.cecs.pdx.edu/~mpj/pubs/instancechains.pdf}.
+
+\bibitem[{Leonardo de Moura} and {Nikolaj Bjørner}(2008)]{z3}
+{Leonardo de Moura} and {Nikolaj Bjørner}.
+\newblock Z3: An efficient smt solver.
+\newblock In \emph{Tools and Algorithms for the Construction and Analysis of
+  Systems, 14th International Conference}, Budapest, Hungary, 2008.
+\newblock URL \url{http://research.microsoft.com/projects/z3/z3.pdf}.
+
+\bibitem[{Mark P. Jones}(1995)]{improvement}
+{Mark P. Jones}.
+\newblock Simplifying and improving qualified types.
+\newblock In \emph{FPCA '95: Conference on Functional Programming Languages and
+  Computer Architecture}, La Jolla, California, USA, 1995.
+\newblock URL \url{http://web.cecs.pdx.edu/~mpj/pubs/fpca95.pdf}.
+
+\bibitem[{Niki Vazou} et~al.(2014){Niki Vazou}, {Eric L. Seidel}, {Ranjit
+  Jhala}, {Dimitrios Vytiniotis}, and {Simon Peyton-jones}]{liquid-haskell}
+{Niki Vazou}, {Eric L. Seidel}, {Ranjit Jhala}, {Dimitrios Vytiniotis}, and
+  {Simon Peyton-jones}.
+\newblock Refinment types for Haskell.
+\newblock In \emph{Proceedings of the 19th ACM SIGPLAN International Conference
+  on Functional Programming}, Gothenburg, Sweeden, 2014.
+\newblock URL
+  \url{http://goto.ucsd.edu/~rjhala/papers/refinement_types_for_haskell.pdf}.
+
+\bibitem[{Patrick C. Hickey} et~al.(2014){Patrick C. Hickey}, {Lee Pike},
+  {Trevor Elliott}, {James Bielman}, and {John Launchbury}]{ivory-experience}
+{Patrick C. Hickey}, {Lee Pike}, {Trevor Elliott}, {James Bielman}, and {John
+  Launchbury}.
+\newblock Building embedded systems with embedded {DSLs} (experience report).
+\newblock In \emph{Proceedings of the 2014 ACM SIGPLAN Conference on Functional
+  Programming}, Gothenburg, Sweden, 2014.
+\newblock URL
+  \url{https://github.com/GaloisInc/smaccmpilot-experiencereport/blob/master/embedded-experience.pdf?raw=true}.
+
+\bibitem[{Richard A. Eisenberg} and {Jan Stolarek}(2014)]{Eisenberg2014}
+{Richard A. Eisenberg} and {Jan Stolarek}.
+\newblock Promoting functions to type families in {Haskell}.
+\newblock In \emph{Proceedings of the 2014 ACM SIGPLAN Symposium on {Haskell}},
+  Gothenburg, Sweden, 2014.
+\newblock URL
+  \url{http://www.cis.upenn.edu/~eir/papers/2014/promotion/promotion.pdf}.
+
+\bibitem[{Richard A. Eisenberg} and {Stephanie Weirich}(2012)]{Eisenberg2012}
+{Richard A. Eisenberg} and {Stephanie Weirich}.
+\newblock Dependently typed programming with singletons.
+\newblock In \emph{Proceedings of the 2012 ACM SIGPLAN Symposium on {Haskell}},
+  Copenhagen, Denmark, 2012.
+
+\bibitem[{Sam Lindley} and {Conor McBride}(2013)]{Lindley2013}
+{Sam Lindley} and {Conor McBride}.
+\newblock Hasochism: The pleasure and pain of dependently typed {Haskell}
+  programming.
+\newblock In \emph{Proceedings of the 2013 ACM SIGPLAN Symposium on {Haskell}},
+  Boston, Massachusetts, USA, 2013.
+\newblock URL
+  \url{https://personal.cis.strath.ac.uk/conor.mcbride/pub/hasochism.pdf}.
+
+\bibitem[{Sergey Berezin} et~al.(2003){Sergey Berezin}, {Vijay Ganesh}, and
+  {Daivd L. Dill}]{cvc4-lin-arith}
+{Sergey Berezin}, {Vijay Ganesh}, and {Daivd L. Dill}.
+\newblock An online proof-producing decision procedure for mixed-integer linear
+  arithmetic.
+\newblock In \emph{Proceedings of the 9th International Conference on Tools and
+  Algorithms for the Construction and Analysis of Systems}, Warsaw, Poland,
+  2003.
+\newblock URL \url{http://hci.stanford.edu/cstr/reports/2007-07.pdf}.
+
+\bibitem[{The GHC Team}(2015)]{ghc-manual}
+{The GHC Team}.
+\newblock \emph{The Glorious Glasgow Haskell Compilation System User's Guide,
+  Version 7.10.1}, 2015.
+\newblock URL
+  \url{https://downloads.haskell.org/~ghc/7.10.1/docs/html/users_guide/index.html}.
+
+\end{thebibliography}
 
